---
title: "POLI 30 D"
author: "Your Name Here"
output: html_document
---

```{r, echo = FALSE, results='hide', include=FALSE}
## Load the tidyverse
library(tidyverse)

## Load the dataset and save it with the name `india`.
india <- read.csv('https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/india.csv')
```

# Problem Set 02 - Do Women Promote Different Policies than Men? (continuation)

## Due: Feb 20

(Based on DSS Materials and on [Chattopadhyay and Esther Duflo. 2004. ``Women as Policy Makers: Evidence from a Randomized Policy Experiment in India."](https://economics.mit.edu/files/792) *Econometrica*, 72 (5): 1409–43.)

We will estimate the average causal effect of having a female politician on two policy outcomes. For this purpose, we will analyze data from an experiment conducted in India, where villages were randomly assigned to have a female council head. The dataset we will use is in a file called "india.csv". The Table below shows the names and descriptions of the variables in this dataset, where the unit of observation is villages.

| **Variable** | **Description**                                                                                  |
|--------------|--------------------------------------------------------------------------------------------------|
| village      | village identifier ("Gram Panchayat number _ village number")                                    |
| female       | whether the village was assigned a female politician: 1=yes, 0=no                                    |
| water        | number of new (or repaired) drinking water facilities in the village <br>since random assignment |
| irrigation   | number of new (or repaired) irrigation facilities in the village<br>since random assignment      |

In this problem set, we will practice loading, making sense of data, and understanding the basics of causal inference. We will also learn how to use R Markdown.

---

### 1. Considering that the dataset we are analyzing comes from a randomized experiment, what can we compute to estimate the average causal effect of having a female politician on the number of new (or repaired) drinking water facilities? Please provide the name of the estimator. (1 point)

**Answer**: Answers here.

---

### 2. In this dataset, what is the average number of new (or repaired) drinking water facilities in villages with a female politician? Please answer with a full sentence. (1 point)

```{r}
## Coding answers here
```

**Answer**: Answers here.

(Hint: we use `[]` to subset a variable; inside the square brackets, we specify the selection criterion. For example, we can use the relational operator `==` to set a logical test; only the observations for which the logical test is true will be extracted.) 


---

### 3. What is the average number of new (or repaired) drinking water facilities in villages with a male politician? Please answer with a full sentence. (1 point)

```{r}
## Coding answers here
```

**Answer**: Answers here.

---

### 4. What is the estimated average causal effect of having a female politician on the number of new (or repaired) drinking water facilities? (2 points)


```{r}
## Coding answers here
```

**Answer**: Put your answers here.

---

### 5. Create a visualization of the distribution of the variable *water*. 
  a. Does this variable look bell-shaped distributed? (0.5 points)
  b. Approximately how many villages in this experiment had about 250 new (or repaired) drinking water facilities since the randomization of politicians? (0.5 points)

```{r}        
## Coding answers here
```

**Answers**: Answers here.

(Hint: the histogram of a variable is the visual representation of its distribution. The function in R to create a histogram is \fu{ggplot(data = dataset, aes(x = x\_variable\_name)) + geom_histogram()}. The only required arguments are the `variable` and the `dataset`.)


---

### 6. Create a visualization of the relationship between *water* and *irrigation*. 
  a. Does the linear relationship between these two variables look positive or negative? A positive/negative answer will suffice. (0.5 points)
  b. Does the relationship between these two variables look strongly linear? A yes/no answer will suffice. (0.5 points) 

```{r}
## Coding answers here
```

**Answers**: Answers here.

(Hint: a scatter plot is the graphical representation of the relationship between two variables. The function in R to create a scatter plot with the fitting line is:

```
ggplot(data = dataset, aes(x = var_x, y = var_y)) + 
  geom_point() + 
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F)
```

It requires three arguments: (1) the name you saved the dataset; (2) the code identifying the variable to be plotted along the x-axis, and (3) the code identifying the variable to be plotted along the y-axis.)

---

### 7. Compute the correlation between *water* and *irrigation*. 
  a. Are you surprised by the sign of the correlation? Provide your reason. (0.5 points)
  b. And are you surprised by the absolute value of the correlation? Provide your reason. (0.5 points)

```{r}
## Coding answers here
```

**Answers**: Answers here.

(Hint: the function in R to compute a correlation coefficient is \fu{cor()}. It requires two arguments (separated by a comma) and in no particular order: the code identifying each of the two variables.) 

---

### 8. If we wanted to use the sample of villages in this dataset to infer the characteristics of all villages in India, we would have to make sure that the sample is _____________ of the population of all villages. (Please provide the missing word). (1 point)

**Answer**: Answers here.

---

### 9. What would have been the best way of selecting the villages for the sample to ensure that the statement above was true? (1 point)

**Answer**: Answers here.