---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 02 | R Basics I}
fontsize: 12pt
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: ../latex/style.tex # changes style to match the style of the book
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

## Before we start

**Announcements:**

- Quizzes and Participation:
  + Start at week 03. We will give full marks on Quiz 1 for everyone on week 03. You're welcome :)

- Github page:

\vspace{0.2cm}

[**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)


## Before we start

**Recap:**

- Last class, you learned how to install R and R Studio on your computer.

**Great job!**

- Do you have any questions about installations? Is it working fine? Let us know!

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 02}}}\normalsize\\\\
- Become familiar with RStudio \\\\
- Become familiar with R \\
\,\,\,\,\,\,\,\,- do calculations: \fu{+}, \fu{-}, \fu{*}, \fu{/}\\
\,\,\,\,\,\,\,\,- create objects: \fu{<-}, \fu{"}\\
\,\,\,\,\,\,\,\,- use functions: \fu{()}, \fu{sqrt()}, \fu{\color{gray}\#\color{text}}\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# Why is it so hard to code?

## If you are worried because...
- You have never done any statistics or coding...
\vspace{.1cm}    
    + Don't worry. We assume you haven't
- Math is your "mortal enemy" or "you are not good at it"...
    + We will start from zero and progress slowly
    + There is plenty of help available
    + As long as you put in the time to do the work every week, you will do great!
- You don't think you will be able to "memorize" all the equations ...
\vspace{.1cm}
    + Start slow and progress. Understand trumps memorization!
\vspace{-1cm}

## Why is it so hard to code?

* Can you read this sentence: "4 exampel in Ingli$h you kin get prackicly evRiThing rong-rong-rong and sti11 be undr3stud."

* You can do it because English is a *Natural language*.

* R, Python, and other programming languages are *formal languages*.

* This means that you should pass the exact instructions to the computer to it do what you want to get done.

* Learning how to code is learning to speak a very formal language. As with learning any language, **practice makes it perfect**!

# Learning your R Studio

## R and RStudio
- In Lab 01, you should have installed two programs on your computer: 
    + R (\includegraphics[height=.05\textheight]{../images/Rlogo.png}) and RStudio (\includegraphics[height=.05\textheight]{../images/RStudio_logo.png})
\vspace{.2cm}   

- \concept{R}\ is the statistical program that will perform calculations and create graphics for us (it's the engine)
\vspace{.2cm}

- \concept{RStudio}\ is the user-friendly interface that we will use to communicate with R
\vspace{.2cm}

- We will never open R directly; we will always start by opening RStudio 
    + RStudio will open R by itself
\vspace{-1cm}

## RStudio
- Go ahead and open RStudio (\includegraphics[height=.05\textheight]{../images/RStudio_logo.png})
\vspace{.2cm}

- Then, open a new R script: 
    + dropdown menu: File > New File > R Script
\vspace{.2cm}

- What is an \concept{R script}? 
    + type of file we use to store the code we write to analyze data    

## RStudio Layout

\begin{center}\includegraphics[height=.65\textheight]{../images/rstudio_windows5}\end{center}


## RStudio Layout

- \concept{R Script}\ (upper left window): where we write and run code
\vspace{.2cm}

- \concept{R Console}\ (lower left window): where R provides the executed code and its outputs, including errors
\vspace{.2cm}

- \concept{Environment}\ (upper right window): storage room of current R session; lists objects that we have created
\vspace{.2cm}

- \concept{Help}\ and \concept{Plots}\ tabs (lower right window)

## We will use R to:

\begin{center}
\begin{tabular}{rl}
\color{text} 1. & Do calculations \\[.3cm]
\color{text} 2. & Create objects \\[.3cm]
\color{text} 3. & Use functions\\
\end{tabular}
\end{center}

## 1. Do calculations
- We can use R as a calculator
\vspace{-.1cm}
    + R understands arithmetic operators such as \fu{+}, \fu{-}, \fu{*}, \fu{/}
\vspace{.2cm}

- Let's ask R to calculate 20 plus 5
\vspace{.2cm}

+ First, we type on the R script (upper left window): \fu{20+5}
\vspace{.2cm}

+ Then, to run this code: we highlight it and either 
\vspace{-.1cm}
    - (a) manually hit the run icon (\includegraphics[height=.5cm]{../images/run.jpg}) or
    - (b) use the shortcut *command+enter* in Mac or *ctrl+enter* in Windows
\vspace{.2cm}

+ Go ahead and do it
\vspace{-1cm}

***       
- In the Console, you should see the following:
\vspace{.2cm}
\begin{center}
\begin{minipage}{0.375\paperwidth}
\begin{Shaded}
\color{blue}{> 20+5}\\
\color{black}{[1] 25}
\end{Shaded}
\end{minipage}
\end{center}
\vspace{.2cm}
    - first, the executed code in \color{blue} blue \color{black} (after \color{blue}>\color{black})
  - then, the output of the executed code in black
  - what does the [1] mean? 
  + it indicates that the output immediately to its right is the first (and only, in this case) output
\vspace{.2cm}

- The title of the R script is now \color{red} red \color{black} to indicate that you have unsaved changes
    + to save the R script either use shortcuts (*command+S* or *ctr+S*) or click on File > Save or Save As...
    + name it "lab02" so that you know what it refers to
    
***
- In the book, as well as in the lectures, we show the output that you should see in the Console right after the code that produces it, like so:
\vspace{.2cm}
\begin{center}
\begin{minipage}{0.375\paperwidth}
\begin{Shaded}
\color{cyan}{20+5}\\
\color{gray}{\#\# [1] 25}
\end{Shaded}
\end{minipage}
\end{center}
\vspace{.2cm}
\vspace{.2cm}
    + \color{cyan}{20+5} \color{black} is the code to be typed and run on the R script (the code that R will execute)
  + the symbol \color{gray}\#\# \color{text} indicates the beginning of the output 
  + \color{gray}{[1] 25} \color{black} is the output (what you should see in the Console after the executed code)

## 2. Create objects
- R stores information in the form of \concept{objects}
- To analyze data, we will need to create objects
- An object is like a box that can contain anything\newline
\vspace{.4cm}
\begin{minipage}{0.4\paperwidth}
\begin{center}\includegraphics[width=4cm]{../images/box3}\end{center}
\end{minipage}\begin{minipage}{0.4\paperwidth}
\color{black}To create one, we need to: 
\begin{itemize}
\item give it a name 
\item specify its contents
\item use the assignment operator \fu{<-}
\end{itemize}
\end{minipage}

***
\color{black} In R, we use the assignment operator \fu{<-} to create an object:
\vspace{.2cm}

- To its left, we specify the name of the object
    + name cannot begin with a number or contain spaces or special symbols like \$ or \% that are reserved for other purposes
    + name can contain \fu{\_} underscores, which are good substitutes for spaces
\vspace{.2cm}

- To its right, we specify the object's content
\vspace{.8cm}
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm]
\begin{center}\fu{\textit{object\_name} <- \textit{object\_content}}\end{center}
\vspace{-.1cm}
\end{mdframed}

***
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm]
\begin{center}\fu{\textit{object\_name} <- \textit{object\_content}}\end{center}
\vspace{-.1cm}
\end{mdframed}
\vspace{.4cm}

- For example, type and run:
    ```{r, collapse=TRUE, comment='##'}
    twentyfive <- 25
    ```
- After running this code, the object *twentyfive* will show up in the Environment (the upper right window of RStudio)

- To find out the contents of an object, you can run the name of the object in R:
    ```{r, collapse=TRUE, comment='##'}
    twentyfive 
    ```

- This is equivalent to asking R: what is inside of *twentyfive*?

***
- Objects can contain text (called strings of characters) as well as numbers

- Run, for example:
    ```{r, collapse=TRUE, comment='##'}
    class <- "POLI30D" 
    ```

- Now, in the environment, there should be two objects
    + What are they? 
\vspace{.2cm}    

- Note that in this last piece of code, we used \fu{"} around the contents, but we did not use \fu{"} in the previous piece of code

\vspace{.1cm}\pause
\begin{center} \fu{twentyfive <- 25} \,\,\,\,\,\,\color{black}vs.\,\,\,\,\,\, \fu{class <- "POLI30D"} \end{center}
- Why?

***
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm]
When do we need to use \fu{"} when writing code in R?

\begin{itemize}
\tightlist
\item the names of objects, names of functions, and names of arguments as well as special values such as TRUE, FALSE, NA, and NULL should NOT be in quotes
\item all other text should be in quotes
\item numbers should never be in quotes unless you want R to treat them as text
\end{itemize}

\vspace{.1cm}
\end{mdframed}    

***
- What would happen if you run instead: \fu{class <- POLI30D}?

-
\vspace{.2cm}
\begin{minipage}{0.7\paperwidth}
\begin{Shaded}
\NormalTok{class <- POLI30D}\\
\color{red}\#\# Error: object 'POLI30D' not found
\end{Shaded}
\end{minipage}
\vspace{.2cm}
    + without the \fu{"}, R thinks that *POLI30D* is the name of an object and R is right; there is no object called *POLI30D* in the environment 
\vspace{.2cm}

- \color{black}**Running into errors is part of the coding process**
    + do not be discouraged
    + if you have problems figuring out what a particular error means, google it; there are lots of Q\&A sites
    + if that doesn't help, post the code and error in our Canvas Discussion

***
- R will overwrite objects if you assign new content to an existing object name
    ```{r, collapse=TRUE, comment='##'}
    class <- "data analysis" 
    ```
    - After running the code above, *class* will contain the text "data analysis" instead of "POLI30D"
\vspace{.2cm}

- R is case sensitive: 
    + *class* is different than *Class*
    + to avoid confusion, we use lower-case letters when naming objects
    
## 3. Use functions
- Think of a function as an action that you request R to perform on a particular object or piece of data, such as calculating the square root of 25

\vspace{.2cm}
\begin{center}
\includegraphics[width=.25\textwidth]{../images/robot.jpg}
\end{center}
\vspace{-1cm}

***
\begin{center}
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm] \vspace{.2cm}
\begin{center}\tip{AN R FUNCTION}\end{center}\color{text}
\begin{center}
\begin{minipage}{0.28\textwidth}\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0.1cm, align=center] \color{text}
\begin{center}takes input(s)\end{center}
\end{mdframed}\end{minipage}
\begin{minipage}{0.05\textwidth}
\color{gray}$\rightarrow$
\end{minipage}
\begin{minipage}{0.28\textwidth}\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0.1cm, align=center] \color{text}
\begin{center}performs action(s) with the input(s)\end{center}
\end{mdframed}\end{minipage}
\begin{minipage}{0.05\textwidth}
\color{gray}$\rightarrow$
\end{minipage}
\begin{minipage}{0.28\textwidth}\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0.1cm, align=center] \color{text}
\begin{center}produces output(s)\end{center}
\end{mdframed}\end{minipage}\end{center}
\vspace{.1cm}
\end{mdframed}
\end{center}
\vspace{.4cm}

- A \concept{function}:
    + takes input(s) 
        + example: takes the number 25
    - performs an action with the input(s) 
        + computes $\sqrt{\textrm{25}}$
    - produces an output 
        + produces the number 5
\vspace{-1cm}

***
- We will learn how to use these functions:
\vspace{.1cm}
    + \fu{sqrt()}, \fu{setwd()}, \fu{read.csv()}, \fu{View()}, \fu{head()}, \fu{dim()}, \fu{mean()}, \fu{ifelse()}, \fu{table()}, 
\fu{prop.table()}, 
<!-- \fu{na.omit()},  -->
\fu{hist()}, \fu{median()}, \fu{sd()}, \fu{var()}, \fu{plot()}, \fu{abline()}, \fu{cor()}, \fu{lm()}, 
<!-- \fu{log()},  -->
\fu{c()}, \fu{sample()}, 
\fu{rnorm()}, \fu{pnorm()}, \fu{print()},
<!-- \fu{nrow()},  -->
<!-- \fu{predict()},  -->
\fu{abs()}, and \fu{summary()}

\vspace{.2cm}
- In time, we will learn:
\vspace{.1cm}
    + their names
    + the actions they perform
    + the inputs they require
    + the outputs they produce

***
- The name of a function (without quotes) is always followed by parentheses: \fu{\textit{function\_name}()}
\vspace{.2cm}

- Inside the parentheses, we specify the inputs, which we refer to as arguments: \fu{\textit{function\_name}(\textit{arguments})}
\vspace{.2cm}

- Most functions require that we specify at least one argument but can take many optional arguments 
    + some arguments are required, others are optional
\vspace{.2cm}

- When multiple arguments are specified inside the parentheses, they are separated by commas: \fu{\textit{function\_name}(\textit{argument1}, \textit{argument2})}
\vspace{.2cm}

***
- To specify the arguments, we enter them in a particular order or include the name of the argument (without quotes) in our specification: 
    + \fu{\textit{function\_name}(\textit{argument1}, \textit{argument2})} or
    + \fu{\textit{function\_name}(\textit{argument1\_name} = \textit{argument1}, \newline $\textrm{\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,}$\textit{argument2\_name} = \textit{argument2})}
\vspace{.1cm}

- We always specify required arguments first
\vspace{.1cm}

- If there is more than one required argument, we enter them in the order expected by R
\vspace{.1cm}

- We specify any optional arguments we want next and include their names:
    + \fu{\textit{function\_name}(\textit{required\_argument},}\newline $\textrm{\,\,\,\,\,\,\,\fu{\textit{optional\_argument\_name} = \textit{optional\_argument})}}$

***
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm] \vspace{.2cm} \color{text}
\begin{center} 
\tip{Using R Functions}: 

We typically write code in one of these two formats:

\vspace{.4cm}
\fu{\textit{function\_name}(\textit{required\_argument})}  \\
\vspace{.2cm}
or \\
\vspace{.2cm}
\fu{\textit{function\_name}(\textit{required\_argument},} \,\,\,\,\,\,\,\fu{\textit{optional\_argument\_name} = \textit{optional\_argument})}
\vspace{.2cm} 
\end{center}
\vspace{-.1cm}
\end{mdframed}    

***
- Fictitious Example: Suppose R were capable of baking and that it had a function named \fu{bake()} that, by default, bakes the specified ingredient for 60 minutes at 400$^{\circ}$F
\vspace{.2cm}
    - Required argument: the ingredient
    \vspace{.1cm}
        + example: cake mix
    - Optional arguments: named \fu{degrees} and \fu{minutes} to change the default temperature and duration of the bake, respectively
    \vspace{.1cm}
        + \fu{degrees=350} changes temperature to 350$^{\circ}$F
        + \fu{minutes=30} changes the duration of the bake to 30 minutes

***
- The following code would ask R to bake a cake mix for 30 minutes at 350$^{\circ}$F, so that we can have cake as the output:

\vspace{.4cm}
\begin{center}
\begin{mdframed}[backgroundcolor=white,linecolor=function, linewidth=0.05cm] \vspace{.2cm}
\begin{center}
\tip{Fictitious Example}\\

\fu{bake( cake\_mix, degrees = 350, minutes = 30 )}
\vspace{.2cm}

\begin{minipage}{0.35\paperwidth}
\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0.1cm, align=center]
\begin{center}\tip{INPUTS}\end{center}
\begin{center}
\color{gray} required argument = \color{text} cake\_mix\\
\color{gray} optional argument \fu{degrees} = \color{text} 350 \\
\color{gray} optional argument \fu{minutes} = \color{text} 30
\end{center}
\vspace{-.1cm}
\end{mdframed}\end{minipage}
\begin{minipage}{0.02\paperwidth}
\begin{center}
\color{gray}$\rightarrow$\\
\end{center}
\end{minipage}
\begin{minipage}{0.18\paperwidth}
\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0cm, align=center]
\tip{ACTION}
\begin{center}
\color{text} bake \\
\end{center}\end{mdframed}\end{minipage}
\begin{minipage}{0.02\paperwidth}
\begin{center}
\color{gray}$\rightarrow$ \\
\end{center}
\end{minipage}
\begin{minipage}{0.19\paperwidth}
\begin{mdframed}[backgroundcolor=shadecolor,linecolor=shadecolor, linewidth=0cm, align=center]
\tip{OUTPUT}
\begin{center}
\color{text} cake\\
\end{center}\end{mdframed}\end{minipage}
\vspace{.2cm}\end{center}\end{mdframed}
\end{center}
\vspace{.5cm}

***
- Example: \fu{sqrt()} computes the square root of the argument specified inside the parentheses. To compute $\sqrt{\textrm{25}}$, run:
    ```{r, collapse=TRUE, comment='##'}
    sqrt(25)
    ```
    + \fu{sqrt} is the name of the function, which, as all function names, is followed by parentheses \fu{()}
    +  \fu{25} is the required argument
    + \color{gray} 5 \color{text} is the output

***
- Alternatively, since the object *twentyfive* contains the number 25, we can run:
    ```{r, collapse=TRUE, comment='##'}
    sqrt(twentyfive)
    ```

\vspace{.2cm}
- R will give you an error message if you run this line of code before creating the object *twentyfive*

- Code is sequential! One must run code in order
    + Whenever returning to work on an R script, run all the code from the beginning
    
***
- It is good practice to comment on code
    + include short notes to yourself or your collaborators explaining what the code does
    
- To comment code, we use \fu{\color{gray}\#\color{black}}
    + R ignores everything that follows this character until the end of the line
    
- Examples:    
    ```{r, collapse=TRUE, comment='##'}
    sqrt(25) # calculates square root of 25
    ```
    ```{r, collapse=TRUE, comment='##'}
    # sqrt(25) calculates square root of 25
    ```

***
- Before closing your computer, remember to save the R script, otherwise, you risk losing unsaved changes
\vspace{.1cm}
    + either use shortcuts (*command+S* or *ctr+S*) or 
    + click on File > Save
\vspace{.2cm}

- If you quit RStudio, R will ask whether you want to save the workspace image, which contains all the objects you have created during the R session
\vspace{.1cm}
    + I recommend that you do not save it
    + You can always re-create the objects by re-running the code in your R script 
    
***
\color{black}
\begin{center}
\begin{tabular}{l}
\concept{\textbf{Today's Lab}}\\
- RStudio: R script, R console, and the environment \\
- R: do calculations: \fu{+}, \fu{-}, \fu{*}, \fu{/}\\
\,\,\,\,\,\,\,\,\,\, create objects: \fu{<-}, \fu{"}\\
\,\,\,\,\,\,\,\,\,\, use functions: \fu{()}, \fu{sqrt()}, \fu{\color{gray}\#\color{text}}\\
\\
\concept{\textbf{Next Lab}}\color{black}\\
- How to load and make sense of data in R\\
- \textbf{Bring your computers!}\\
\end{tabular}
\end{center}
\vspace{-1cm}
  
# Questions?

# See you in the next lab!