---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 11 | Causal Inference with Observational Data II}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- GitHub page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: Not sure what the link is. Ask your TA!

- Note to self: Turn on the mic!

## Before we start

**Recap:** We learned:

- The definitions of theory, scientific theory, and hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.
- Sampling, descriptive statistics, and descriptive plots for one variable.
- Correlation between two continuous variables.
- Prediction of a binary and a non-binary variable.
- ATE using differences-in-means and simple linear regression.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.80\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\[.3cm]
- How Can We Estimate Causal Effects with\\
\,\,\,\,\,Observational Data?\\\\
- Multiple Linear Regression Models \\
\,\,\,\,\, - Interpretation of Coefficients\\
\,\,\,\,\, - Interpretation of $\widehat{\beta_1}$ When X$_1$ Is the Treatment\\
\,\,\,\,\,\,\,\,\,\, Variable and the Other X Variables Are All\\
\,\,\,\,\,\,\,\,\,\, the Potential Confounding Variables\\\\
- What is the Effect of the Death of the Leader \\
\,\,\,\,\,on the Level of Democracy?\\
\end{tabular}
\end{center}
\vspace{.1cm}
\end{mdframed}
\end{minipage}
\end{center}

##  How Can We Estimate Causal Effects\newline with Observational Data?

- Most of the time, we have no control over the assignment.
  + Then, how to eliminate potential confounders and make treatment and control groups comparable?

\vspace{.1cm}

- First, we must identify all potentially confounding variables
  + variables that affect both (i) the likelihood of receiving the treatment and (ii) the outcome

\vspace{.2cm}
\begin{center}
\begin{tikzpicture}
\node (x) at (1,1) {$X$};
\node (z) at (2,2) {$Z$};
\node (y) at (3,1) {$Y$};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (z) -- (x);
\end{tikzpicture}
\end{center}
\vspace{.1cm}

- Then, we need to *control* for them by fitting a **multiple linear regression model**

\vspace{-1cm}

## Multiple Linear Regression Models

Linear models with more than one $X$ variable

\vspace{-.3cm}

$$\widehat{Y}_i = \widehat{\beta}_0 + \widehat{\beta}_1 X_{i1}+... + \widehat{\beta}_p X_{ip} $$
\vspace{-.8cm}

\noindent where:
\begin{itemize}
\item $\widehat{Y_i}$ is the predicted value of $Y$ for observation $i$
\vspace{-.1cm}
\item $\widehat{\beta}_0$ is the estimated intercept coefficient  
\vspace{-.1cm}
\item each $\widehat{\beta}_j$ (pronounced `beta hat sub j`) is the estimated coefficient for variable $X_j$ ($j \in \{1, ..., p\}$)
\vspace{-.1cm}
\item each $X_{ij}$ is the observed value of the variable $X_j$ for observation $i$ ($j \in \{1, ..., p\}$ and $i \in \{1, ..., n\}$)
\vspace{-.1cm}
\item $p$ is the total number of $X$ variables in the model.
\end{itemize}
\vspace{-1cm}

## Multiple Linear Regression Models

\begin{table}[h!]
\begin{center}
\begin{tabular}{@{\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,}c@{\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,}c@{\,\,\,\,\,\,\,\,\,\,\,\,\,\,}}
\arrayrulecolor{gray}\midrule
 simple regression & multiple regression\\\addlinespace[2.5pt]
$\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$ & $\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_{1} X_{1}+ ... + \widehat{\beta}_p X_{p}$
\\\addlinespace[3pt]
\arrayrulecolor{gray}\midrule \addlinespace[3pt]
$\widehat{\beta}_0$: $\widehat{Y}$ when $X{=}{0}$ &$\widehat{\beta}_0$: $\widehat{Y}$ when all $X_j{=}{0}$ \\
& ($j{=}{1},...,p$) \\\addlinespace[4pt]
$\widehat{\beta}_1$: $\triangle\widehat{Y}$ associated & each $\widehat{\beta}_j$: $\triangle\widehat{Y}$ associated \\
with $\triangle X{=}{1}$ & with $\triangle X_j{=}{1}$, \\ 
& while holding all other \\
& $X$ variables constant \\
& or \textit{ceteris paribus}\\\addlinespace[2pt]
\arrayrulecolor{gray}\midrule
\end{tabular}
\end{center}
\vspace{-.2cm}
\end{table}

## Interpretation of Coefficients in Multiple Linear Regression Models

- $\widehat{\beta}_0$ is the $\widehat{Y}$ when \textit{all} $X_j{=}{0}$

\vspace{.2cm}

- Because there are multiple $X$ variables, there are multiple $\widehat{\beta}_j$ coefficients (one for each $X$ variable)

\vspace{.2cm}

- Each $\widehat{\beta}_j$ is the $\triangle \widehat{Y}$ associated with $\triangle X_j=1$, \textbf{while holding all other $X$ variables constant}

\vspace{-1cm}

## Interpretation of Coefficients in Multiple Linear Regression Models

- Adding all confounders as controls in the model makes treatment and control groups comparable **after controls**:
  + Something fancily called `conditional exogeneity`.

- As a result, if $\widehat{\beta}_1$ is the treatment of interest, we can interpret it using the \concept{causal language}.

+ $\widehat{\beta}_1$ is the $\triangle \widehat{Y}$ \textit{caused by} the presence of the treatment ($\triangle X_1$=1), while holding confounders constant.

- $\widehat{\beta}_1$ should be a valid estimate of the average treatment effect if **all** confounding variables are controlled.

- Is this even possible?

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

\vspace{.6cm}\begin{center}\includegraphics[height=.5\textheight]{../../images/leaders}\end{center}

\footnotesize

(Based on \href{https://economics.mit.edu/files/3055}{Jones and Olken. 2009. *Hit or Miss? The Effect of Assassinations on Institutions and War.* \textbf{American Economic Journal: Macroeconomics}, 1 (2): 55-87.})

## Does the Death of the Leader Affect the Level of Democracy?

- After an assassination attempt, the death of a leader is close to random:
  + Leaders whose assassination attempts succeeded should be, on average, *comparable* to leaders whose attempts failed.
  + Do you believe that?

- If this is true, we can estimate the average causal effect of the leader's death.

- Fitting a simple linear model where the death of a leader is the "treatment" variable.

\vspace{-1cm}

## The *leaders* dataset

- Dataset on assassinations and assassination attempts against political leaders from 1875 to 2004.

\footnotesize
| **variable** | **meaning** |
|---|---|
| year | year of the assassination attempt |
| country | country where the assassination attempt took place |
| leadername | name of the leader |
| died | whether the leader died as a result of the assassination attempt |
| politybefore | [Polity scores](https://en.wikipedia.org/wiki/Polity_data_series) of the country before the assassination attempt |
| polityafter | [Polity scores](https://en.wikipedia.org/wiki/Polity_data_series) of the country after the assassination attempt |

\small

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

1. Open RStudio
\vspace{.2cm}

2. Open **inclass07.R** from within RStudio. The file is on Canvas and the [Class GitHub](https://github.com/umbertomig/POLI30Dpublic).
\vspace{.2cm}

3. Load the dataset and check the first observations.

\vspace{-1cm}

## What is the Effect of the Death of the Leader on the Level of Democracy?

\footnotesize
```{r, collapse=TRUE, comment='##', mytiny = TRUE}
leaders <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/leaders.csv")
head(leaders)
```

\normalsize

- What is the Treatment variable (X)?

- What is the Outcome variable (Y)?

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

- To fit the simple linear model, where $\widehat{\beta}_1$ is equivalent to the difference-in-means estimator, we run:

```{r, collapse=TRUE, comment='##', myfntsize = TRUE, fig.width=8, fig.height=5}
lm(polityafter ~ died, data = leaders)  
```

- Fitted model: $\widehat{\textrm{\textit{polityafter}}}$ = -1.90 \pause + 1.13 *died*

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$:**

+ $\widehat{\beta}_1$ is the $\triangle \widehat{Y}$ associated with $\triangle X$ = 1

+ Here: $\widehat{\beta}_1$ = 1.13 is the $\triangle \widehat{\textrm{\textit{polityafter}}}$ associated with $\triangle$*died* = 1

+ In words: the death of the leader is associated with a predicted increase in polity scores of 1.13 points, on average

## Does the Death of the Leader Affect the Level of Democracy?

**Unit of measurement of $\widehat{\beta}_1$**:

+ Same as $\triangle \overline{Y}$

+ In here, *Y* is non-binary and measured in points.

+ Then, as $\triangle \overline{Y}$ is measured in points, so is $\widehat{\beta}_1$

\vspace{-1cm}
    
## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$ (continuation):**

- *died* in the attack is the "treatment" variable.

- The Polity IV score is the "outcome" of interest.
  + Thus, $\widehat{\beta}_1$ is equivalent to the difference-in-means estimator
  
- We may interpret $\widehat{\beta}_1$ using \concept{causal language}:

\vspace{0.2cm}

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.1cm}
\noindent The death of the leader is associated with a predicted \concept{increase}\ in polity scores after the assassination by 1.13 points, on average.
\vspace{.1cm} 
\end{mdframed}

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$ (continuation):**

- Assumption for Causal Claim be valid:

  + Assassination attempts where the leader died are comparable to those where the leader did not die

- Again, is this plausible? 
  + Can you think of a confounder?

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

- We can look at *pre-treatment characteristics*: how democratic was the country *before* the assassination attempt?

- Calculate the average *politybefore* for the two groups:

```{r, collapse=TRUE, comment='##', mysmall = TRUE}
mean(leaders$politybefore[leaders$died==1]) # treatment
mean(leaders$politybefore[leaders$died==0]) # control
```

- What do you see in here?

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

- Attempts where the leader died were more democratic to begin with.

- Thus, *politybefore* might be a **confounding variable**:

\vspace{0.2cm}

\begin{center}
\begin{tikzpicture}
\node (x) at (0,1) {\textit{died}};
\node (z) at (2,2) {\textit{politybefore}};
\node (y) at (4,1) {\textit{polityafter}};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (z) -- (x);
\end{tikzpicture}
\end{center}

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

**Estimate average causal effect while controlling for confounders**:

- To control for confounders, we need to fit the following multiple regression linear model:

\begin{center}$\widehat{\textrm{\textit{polityafter}}}$ = $\widehat{\beta}_0$ + $\widehat{\beta}_1$ \textit{died} + $\widehat{\beta}_2$ \textit{politybefore}\end{center}
\vspace{.2cm}

- To fit the model, we use the function \fu{lm()}\ with a formula of the type \fu{Y $\sim$ X$_1$ + X$_2$}

## Does the Death of the Leader Affect the Level of Democracy?

```{r, collapse=TRUE, comment='##', mysmall = TRUE}
lm(polityafter ~ died + politybefore, data=leaders)  
```

- Fitted model:

\vspace{.2cm}
\begin{center}
$\widehat{\textrm{\textit{polityafter}}}$ = -0.43 + 0.26 \textit{died} + 0.84 \textit{politybefore}
\end{center}

## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$:**

- $\widehat{\beta}_1$ is the $\triangle \widehat{Y}$ associated with $\triangle X_1$=1, while holding all other variables constant.

  + Here: $\widehat{\beta}_1$ = 0.26 is the $\triangle \widehat{\textrm{polityafter}}$ associated with $\triangle$*died*=1, while holding *politybefore* constant.

\vspace{0.2cm}

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.1cm}
\noindent The death of the leader is associated with a predicted increase in polity scores after the assassination attempt of 0.26 points, on average, while holding polity scores before constant.
\vspace{.1cm} 
\end{mdframed}

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$:**

- Unit of measurement of $\widehat{\beta}_1$

  + Same as $\triangle \overline{Y}$
  
  + Here, *Y* is non-binary and measured in points so $\triangle \overline{Y}$ is measured in points, and so is $\widehat{\beta}_1$

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

**Interpretation of $\widehat{\beta}_1$ (continuation)**

\vspace{.1cm}

- Since here $X_1$ is the treatment variable, $Y$ is the outcome variable of interest, and $X_2$ is the confounder we are worried about, we can interpret $\widehat{\beta}_1$ using \concept{causal language}

- **Causal language:** We estimate that the death of the leader \concept{\textit{increases}}\ polity scores after the assassination attempt by 0.26 points, on average, when holding polity scores before the assassination attempt constant.

\vspace{-1cm}

## Does the Death of the Leader Affect the Level of Democracy?

- This should be a valid estimate of the average treatment effect if *politybefore* is the only confounder

- Note that once we control for *politybefore*, the effect size decreases substantially (it goes from 1.13 to 0.26)

- Based on this analysis, the death of the leader increases the level of democracy of a country but by a smaller amount (more on this later)

\vspace{-1cm}

***
\linespread{1.5}\selectfont
\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.2cm}
\noindent \tip{Average Causal Effects with Observational Data and Multiple Linear Regression}.\\Suppose in the multiple linear regression model where $X_1$ is the treatment variable, we control for \textit{all} potential confounders by including them in the model as additional $X$ variables. In that case, we can interpret $\widehat{\beta}_1$ as a valid estimate of the average causal effect of $X_1$ on $Y$.
\vspace{.2cm} 
\end{mdframed}

## Summary

- **Today's Class:**
  + With Observational Data, Use Multiple Linear Regression Models to:
    1. Control for Confounders
    2. Estimate Average Treatment Effects

\vspace{0.2cm}

- **Next class**:
  + More Causality with Observational Data:
    + Internal versus External Validity.
  
\vspace{-1cm}

# Questions?

# See you in the next class!