---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 03 | R Basics II}
fontsize: 12pt
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r echo=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Github page: [**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: [https://piazza.com/ucsd/winter2023/17221](https://piazza.com/ucsd/winter2023/17221)

## Before we start

**Recap:** In the Lab sessions, you learned:

- How to install R and R Studio on your computer.
- How to do basic math operations in R.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 03}}}\normalsize\\\\
- Learn R data types\\
- Learn how to create datasets\\
- Learn how to load a dataset\\
- Learn how to explore a dataset\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# R Data Types

## R Numeric and R Character

You can create numeric and character variables in R easily:

```{r}
# Numeric
x <- 20
# Character (or string, same thing!)
y <- 'POLI 30 D'
print(x)
print(y)
```

## R Numeric and R Character

We can easily check the types with the \fu{class(.)}  function:

```{r}
class(x)
class(y)
```

## R Numeric and R Character

The \fu{str(.)} function also gives a neat description of what is going on. 

- `str` is short for the structure of the data:

```{r}
str(x)
str(y)
```

## R Vectors

To create a vector, you use the function \fu{c(.)}, and separate the values with a comma:

```{r, mysmall = TRUE}
voted   <- c(1, 0, 0, 1) # Binary vector (1=yes; 0=no)
age     <- c(48, 23, 18, 33) # Numeric vector with ages
message <- c('yes', 'yes', 'no', 'no') # Got message?
place   <- c('La Jolla', 'Del Mar', 
             'Del Mar', 'Poway') # Character with places
```

Each of the variables has four positions (like four observations).

## R Vectors

We can check the data for each of the four observations we have. We use square brackets (\fu{[} and \fu{]}) to index.

First person in the `voted` variable:

```{r, mysmall = TRUE}
voted[1]
```

\vspace{0.3cm}

Ages of first three people. Note `1:3` (try it on the console!). It creates sequences.

```{r, mysmall = TRUE}
age[1:3]
```

## R Vectors

Messages for all people but the second person:

```{r, mysmall = TRUE}
message[-2]
```

Checking all the places vector:

```{r, mysmall = TRUE}
place
```

Changing the age of the second person:

```{r, mysmall = TRUE}
age[2] <- 24
age
```

## R Vectors

You can also create a numeric vector from a character vector. You should use the function \fu{ifelse(.)}:

```{r, mysmall = TRUE}
message
message_num <- 
  ifelse(message == 'yes', # Check if message is yes
         1, # change to 1 if yes
         0) # change to 0 if no
message_num
```

## R Vectors

You can also check the length of the vector:

```{r, mysmall = TRUE}
length(message)
```

\vspace{0.3cm}

And if you don't need an object anymore, you can remove it:

```{r, mysmall = TRUE}
rm(message_num)
```

\vspace{0.3cm}

Now check the environment. `message_num` should have disappeared!

# R `data.frame`

## Creating a `data.frame` from scratch

To create a \fu{data.frame(.)} from scratch, you can just add variables inside it:

```{r, mysmall = TRUE}
dat <- data.frame(
  v1 = c(1,2,3),
  v2 = c('a', 'b', 'c'),
  v3 = c('Treatment', 'Control', 'Control')
)
dat
```

## Creating a `data.frame` from scratch

**Your turn**: Create a data frame with the following info and call it `dat2`:

\begin{center}\includegraphics[height=.45\textheight]{../../images/dat1.png}\end{center}

## Creating a `data.frame` from scratch

If you are curious, these are the meanings:

| **Variable** | **Meaning**                                                                                          |
|--------------|------------------------------------------------------------------------------------------------------|
| age          | Age in years                                                                                         |
| college      | Yes means college complete                                                                           |
| voted        | 1 means voted                                                                                        |
| work         | FT means full-time worker; PT means partial-time worker; UN means unemployed; RT means retired |

## Creating a `data.frame` from existing variables

If you recall, we created the following variables: `voted`, `age`, `message`, and `place`.

Here is how to create a data.frame with them:

```{r, mysmall = TRUE}
dat3 <- data.frame(voted, age, message, place)
dat3
```

# Loading CSV data in R

## Loading a CSV dataset from the locale

- CSV stands for [Comma Separated Values](https://en.wikipedia.org/wiki/Comma-separated_values). It is a particular way to organize data:
  + Each line corresponds to one observation
  + Within lines, information for each variable is separated by a comma.

- To load data from the locale (i.e., your computer), you must find and change your working directory.

- The book explains more about that. We are not going to deal with these cases here.

- We will frequently analyze data from **GitHub**.

## Loading a CSV dataset from GitHub

To load a CSV dataset from GitHub:

1. [Open the GitHub](https://github.com/umbertomig/POLI30Dpublic)
2. [Go to the GitHub data folder](https://github.com/umbertomig/POLI30Dpublic/tree/main/data)
3. [Select the Dataset you want to open (in the case, `countries.csv`)](https://github.com/umbertomig/POLI30Dpublic/blob/main/data/countries.csv)
4. [Find the *Raw* Version of it](https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/countries.csv)
5. Copy the URL (CMD + C / Ctrl + C). In the case of countries: https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/countries.csv
6. Do `name_data_frame <- read.csv('paste_URL_in_here')`

## Example: `countries.csv`

\begin{center}\includegraphics[height=.7\textheight]{../../images/gdata.png}\end{center}

## Example: `countries.csv`

\begin{center}\includegraphics[height=.7\textheight]{../../images/ghcclickraw.png}\end{center}

## Example: `countries.csv`

\begin{center}\includegraphics[height=.4\textheight]{../../images/ghcraw1.png}\end{center}

\begin{center}\includegraphics[height=.2\textheight]{../../images/ghcraw2.png}\end{center}

## Example: `countries.csv`

And since the link is: `https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/countries.csv`:

```{r, mysmall = TRUE}
countries <- read.csv('https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/countries.csv')
head(countries)
```

# Exploring a dataset

## `head` and `tail`

\fu{head(.)} shows the first six observations:

```{r, myfntsize = TRUE}
head(countries)
```

## `head` and `tail`

\fu{tail(.)} shows the last six observations:

```{r, myfntsize = TRUE}
tail(countries)
```

## `str`

That is a nice and neat description of the data.frame. You can try it with other objects too!

```{r, myfntsize = TRUE}
str(countries)
```

## `summary`, `View`, and `dim`

- \fu{dim(.)} gives you the dimension of the data.frame. 
  + The first number is the number of rows (observations).
  + The second number is the number of columns (variables).

```{r}
dim(countries)
```

- The \fu{View(.)} function is also useful. In your console, type the following:

```
View(countries)
```

- Very cool, right?!

## `summary`, `View`, and `dim`

And for a quick summary of the variables in your dataset, you can use the function \fu{summary(.)}:

\vspace{0.2cm}

```{r, mytiny = TRUE}
summary(countries)
```

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Today's Lab}}}\normalsize\\
- Learn R data types\\
- Learn how to create datasets\\
- Learn how to load a dataset\\
- Learn how to explore a dataset\\\\
\multicolumn{1}{c}{\large\concept{\textbf{Next Lab}}}\normalsize\\
- How to operate with data.frames\\
- Learn how to use R Markdown\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}
  
# Questions?

# See you in the next lab!