---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 10 | Causal Inference with Observational Data I}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- GitHub page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: Not sure what the link is. Ask your TA!

- Note to self: Turn on the mic!

## Before we start

**Recap:** We learned:

- The definitions of theory, scientific theory, and hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.
- Sampling, descriptive statistics, and descriptive plots for one variable.
- Correlation between two continuous variables.
- Prediction of a binary and a non-binary variable.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.80\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\[.3cm]
- Review: Causation and Randomized Experiments\\[.3cm]
- Observational Studies\\[.3cm]
- Confounding Variables or Confounders\\
\,\,\,\,\,- Why Are Confounders a Problem?\\
\,\,\,\,\,- Why Don't We Worry About Confounders\\
\,\,\,\,\,\,\,\,\,in Randomized Experiments?\\[.3cm]
- How Can We Estimate Causal Effects with\\
\,\,\,\,\,Observational Data?\\
\,\,\,\,\,- Interpretation of $\widehat{\beta}$ When X Is the Treatment\\
\,\,\,\,\,\,\,\,\,Variable and Y Is the Outcome Variable\\
\end{tabular}
\end{center}
\vspace{.1cm}
\end{mdframed}
\end{minipage}
\end{center}

## Review: Causation

- To measure causal effects, we need to compare the factual outcome with the counterfactual outcome
    + Fundamental problem: We can never observe counterfactual outcomes

- To estimate causal effects, we must find or create a situation in which the treatment and control groups are \concept{comparable}.

- Only when that assumption is satisfied can we use the *factual* outcome of one group as a good *proxy* for the *counterfactual* outcome of the other.

\vspace{-1cm}

## Review: Randomized Experiments

- In randomized experiments, we can rely on the \concept{random assignment of treatment}\ to make treatment and control groups, on average, identical

- Thus, we can estimate the average treatment effect with the \concept{difference-in-means estimator}

$$\overline{Y}_\text{treatment group} - \overline{Y}_\text{control group}$$

## Observational Data

- But what happens when we cannot conduct a randomized experiment and have to analyze observational data?
  + *Observational data*: data collected about naturally occurring events (i.e., researchers do not get to assign the treatment)

- We can no longer assume that treatment and control groups are comparable.

- We have to identify any relevant differences between treatment and control groups (known as confounding variables or confounders)

- Then, we have to statistically control for them so that we may claim that the two groups are comparable.
\vspace{-1cm}

## Confounders or Confounding Variables

- A \concept{confounding variable}, or \concept{confounder}, is a variable that affects both:
    1. The likelihood of receiving the treatment $X$, and
    2. The outcome $Y$

\vspace{.2cm}

- In mathematical notation, we represent a potential confounding variable as $Z$

\vspace{.2cm}
\begin{center}
\begin{tikzpicture}
\node (x) at (1,1) {$X$};
\node (z) at (2,2) {$Z$};
\node (y) at (3,1) {$Y$};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (z) -- (x);
\end{tikzpicture}
\end{center}

## Confounders or Confounding Variables

- Suppose we are interested in the average causal effect of attending a private school instead of a public one on SAT performance.

\vspace{.2cm}
  + What is the treatment variable $X$?
  + What is the outcome variable $Y$?
  + Can you think of a potential confounder $Z$?

\vspace{.2cm}
\begin{center}
\begin{tikzpicture}
\node (x) at (0.5,0) {\textit{private} \textit{school}};
\node (z) at (2,1) {\textit{family} \textit{wealth}};
\node (y) at (3.5,0) {\textit{test} \textit{scores}};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (z) -- (x);
\end{tikzpicture}
\end{center}

## Why Are Confounders a Problem?

- They obscure the causal relationship between $X$ and $Y$!

- In the example above, if we observed that, on average, private school students perform better than public school students, we would not know whether it is:
  + Because they attended a private school or 
  + Because they came from wealthier families
    
- We would not know what portion of the observed differences in SAT performance (the difference-in-means estimator), if any, could be attributed to:
  + Attending a private school versus
  + Coming from a wealthy family.

\vspace{-1cm}

## Why Are Confounders a Problem?

\begin{center}
\includegraphics[height=.80\textheight]{../../images/correlation_not_causation.jpg}
\end{center}

\vspace{-2cm}

## Why Are Confounders a Problem?

- In the presence of confounders, correlation does not necessarily imply causation.

- Just because two variables are highly correlated, it does **not** mean that one causes the other:
    + There could be a third variable that causes both!

- Ice cream sales and shark attacks are highly correlated. 
  + Does this mean that eating ice cream increases the probability that a shark will attack you?

\begin{center}
\begin{tikzpicture}
\node (x) at (0,0) {\textit{ice} \textit{cream} \textit{sales}};
\node (z) at (2,1) {\textit{heat}};
\node (y) at (4,0) {\textit{shark} \textit{attack}};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (z) -- (x);
\end{tikzpicture}
\end{center}

\vspace{-2cm}

## Why Are Confounders a Problem?

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.2cm}
\tip{\concept{In the presence of confounders}}

- Correlation does not imply causation.

- The treatment and control groups are not \textbf{comparable}.

- The difference-in-means estimator does \textbf{NOT} provide a valid estimate of the average causal effect!

\vspace{.2cm}
\end{mdframed}

\vspace{-1cm}

## Why Don't We Worry About Confounders in Randomized Experiments?

- Randomization of treatment assignment eliminates all potential confounders.
  + That is this is the \color{yellow}gold standard\color{black}\ for causal inference.

- Ensures that treatment and control are comparable by breaking the link between any potential confounder.

- If we have a lottery to randomly determine who will attend the private school, then we break the wealth link.

\vspace{.1cm}\pause
\begin{center}
\begin{tikzpicture}
\node (r) at (-3,0) {\textrm{\textit{lottery}}};
\node (x) at (0,0) {\textrm{\textit{private school}}};
\node (z) at (1.5,1) {\textrm{\textit{family wealth}}};
\node (y) at (3,0) {\textrm{\textit{test scores}}};
\draw[->, line width=.02cm] (z) -- (y);
\draw[->, line width=.02cm] (x) -- (y);
\draw[->, line width=.02cm] (r) -- (x);
\end{tikzpicture}
\end{center}
\vspace{-1cm}

##  How Can We Estimate Causal Effects\newline with Observational Data?

- We cannot rely on random treatment assignments to eliminate potential confounders.

- We must identify all potential confounders and statistically control for them using a multiple linear regression model.

- Before we learn how to do that, we will fit a linear regression model to find the *difference-in-means estimator*.

\vspace{-1cm}

## Using the Linear Regression to Compute the Difference-in-Means Estimator

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.2cm} \color{text}
When $X$ is the treatment variable, and $Y$ is the outcome variable of interest, the estimated slope coefficient ($\widehat{\beta}_1$) is \textbf{equivalent} to the \textit{difference-in-means estimator}.
\vspace{.2cm}
\end{mdframed}

- Let us return to a beloved example: *Does Social Pressure Affect Turnout?*

- Registered voters were randomly assigned to either:
  a. receive a message designed to induce social pressure or 
  b. receive nothing

\vspace{-2cm}

## Does Social Pressure Affect Turnout?
\vspace{.4cm}\begin{center}\includegraphics[height=.6\textheight]{../../images/vote.jpg}\end{center}\vspace{-.3cm}

\footnotesize
(Based on \href{http://www.donaldgreen.com/wp-content/uploads/2015/09/Gerber_Green_Larimer-APSR-2008.pdf}{Alan S. Gerber, Donald P. Green, and Christopher W. Larimer. 2008. ``Social Pressure and Voter Turnout: Evidence from a Large-Scale Field Experiment." \textit{American Political Science Review}, 102 (1): 33-48.})

## Does Social Pressure Affect Turnout?
1. Load and look at the data:

```{r, collapse=TRUE, comment='##', echo=T}
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv") # loads and stores data
head(voting) # shows first six observations
```

2. Creating the treatment variable:

```{r, collapse=TRUE, comment='##', echo=T}
voting$pressure <-  ifelse(voting$message=="yes", 1, 0)
```

## Does Social Pressure Affect Turnout?

3. Compute difference-in-means estimator directly

```{r, collapse=TRUE, comment='##'}
mean(voting$voted[voting$pressure==1]) -
  mean(voting$voted[voting$pressure==0]) 
```

4. Alternatively, we can fit a linear model where X is the treatment variable and Y is the outcome variable.

## Does Social Pressure Affect Turnout?

- Recall: the R function to fit a linear model is \fu{lm()}

```{r, collapse=TRUE, comment='##', myfntsize = TRUE}
lm(voted ~ pressure, data=voting) 
```

- Fitted model: $\widehat{\textrm{\textit{voted}}}$ = 0.30 + 0.08 *pressure*

- Note that $\widehat{\beta}_1$ has the same value as the difference-in-means estimator above (both equal 0.08)


## Interpretation of $\widehat{\beta}_1$ When X Is the Treatment Variable, and Y Is the Outcome Variable

- Start the same as in predictive models:
  + Definition: $\widehat{\beta}$ is the $\triangle \widehat{Y}$ associated with $\triangle X$=1
    + $\widehat{\beta}$ = 0.08 is the $\triangle \widehat{\textrm{\textit{voted}}}$ associated with $\triangle$*pressure*=1
  + Receiving a social-pressure inducing message is associated with a predicted increase in the probability of voting of 8 p.p., on average

- Unit of measurement of $\widehat{\beta}_1$? same as $\triangle \overline{Y}$.
  + Since *Y* is binary, $\triangle \overline{Y}$ is measured in p.p., and so is $\widehat{\beta}$ (after x 100)

\vspace{-1cm}
    
## Interpretation of $\widehat{\beta}_1$ When X Is the Treatment Variable, and Y Is the Outcome Variable

- Since $X$ is the treatment variable and $Y$ is the outcome variable, $\widehat{\beta}_1$ is equivalent to the difference-in-means estimator

- As a result, we can interpret $\widehat{\beta}_1$ using \concept{causal langauge}

- **Predictive language**: We estimate that receiving the message inducing social pressure \concept{\textit{is associated with a predicted increase}}\ in the probability of voting of 8 p.p., on average

- **Causal language**: We estimate that receiving the message inducing social pressure \concept{\textit{increases}}\ the probability of voting by 8 p.p., on average

\vspace{-1cm}

## Interpretation of $\widehat{\beta}_1$ When X Is the Treatment Variable, and Y Is the Outcome Variable

- This should be a valid estimate of the average treatment effect if there are no confounding variables present:

  + If registered voters who received the message are comparable to those who did not.

- Since the data come from a randomized experiment, there should be no confounding variables (why?)

- And thus, the difference-in-means estimator should produce a valid estimate of the average treatment effect

\vspace{-1cm}

## Interpretation of $\widehat{\beta}_1$ When X Is the Treatment Variable, and Y Is the Outcome Variable

- **Conclusion:** A message inducing social pressure increases the probability of voting by eight p.p., on average.
  + Valid estimate of the ATE if registered voters who received the message are comparable to those who did not. 
  + This is a reasonable assumption, given that the data come from a randomized experiment.

- Note that this is the same conclusion we arrived at in a previous lecture.

\normalsize

## Interpretation of $\widehat{\beta}_1$ When X Is the Treatment Variable, and Y Is the Outcome Variable

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.2cm} \color{text}
\noindent \tip{Interpretation of the Estimated Slope \newline Coefficient in the Simple Linear Model}: 
\begin{itemize}
\item By default, we interpret $\widehat{\beta}_1$ using predictive language: It is the $\triangle \widehat{Y}$ \textit{associated with} $\triangle X{=}\textrm{1}$.
\item When $X$ is the treatment variable, then $\widehat{\beta}_1$ is equivalent to the difference-in-means estimator and, thus, we interpret $\widehat{\beta}_1$ using causal language: It is the $\triangle \widehat{Y}$ \textit{caused by} $\triangle X{=}\textrm{1}$. This causal interpretation is valid if no confounding variables exist: the treatment and control groups are comparable.
\end{itemize}
\vspace{.2cm}
\end{mdframed}

\vspace{-2cm}

## Summary

- **Today's Class:**
  + Observational Studies
  + Confounding Variables or Confounders
    + Why Are Confounders a Problem?
    + Why Don't We Worry About Confounders in Randomized Experiments?
  + How Can We Estimate Causal Effects with Observational Data?
  + Interpretation of $\widehat{\beta}_1$ when X is the Treatment Variable, and Y Is the Outcome Variable.

\vspace{0.2cm}

- **Next class**:
  + More Causality with Observational Data:
    + We will use *Multiple Regression* models to control for confounders.
  
\vspace{-1cm}

# Questions?

# See you in the next class!