---
title: "POLI 30 D"
author: "Your name here"
output: html_document
---

# Problem Set 01 - Do Women Promote Different Policies than Men?

(Based on DSS Materials and on [Chattopadhyay and Esther Duflo. 2004. ``Women as Policy Makers: Evidence from a Randomized Policy Experiment in India."](https://economics.mit.edu/files/792) *Econometrica*, 72 (5): 1409–43.)

We will estimate the average causal effect of having a female politician on two policy outcomes. For this purpose, we will analyze data from an experiment conducted in India, where villages were randomly assigned to have a female council head. The dataset we will use is in a file called "india.csv". The Table below shows the names and descriptions of the variables in this dataset, where the unit of observation is villages.

| **Variable** | **Description**                                                                                  |
|--------------|--------------------------------------------------------------------------------------------------|
| village      | village identifier ("Gram Panchayat number _ village number")                                    |
| female       | whether village was assigned a female politician: 1=yes, 0=no                                    |
| water        | number of new (or repaired) drinking water facilities in the village <br>since random assignment |
| irrigation   | number of new (or repaired) irrigation facilities in the village<br>since random assignment      |

In this problem set, we will practice loading, making sense of data, and understanding the basics of causal inference. We will also learn how to use R Markdown.

---

1. Use the function `read.csv()` to read the CSV file "india.csv". You can find it on the GitHub page: [https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/india.csv](https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/india.csv). Use the assignment operator `<-` to store the data in an object called *india*. Provide the R code you used. (1 point) 


```{r}
# Your coding answers here.
```

---
    
2. Use the function `head()` to view the first few observations of the dataset. Provide the R code you used. (1 point)

```{r}
# Your coding answers here.
```

---

3. What does each observation in this dataset represent? Please substantively interpret the first observation in the dataset. (1 point)

**Answer:** Change this line to type your answer here.

---

4. For each variable in the dataset, please identify the type of variable (character vs. numeric binary vs. numeric non-binary) (1 point)

**Answer:** Change this line to type your answer here.

---

5. How many observations are in the dataset? In other words, how many villages were part of this experiment? (Hint: the function `dim()` might be helpful here.) (1 point)

```{r}
# Your coding answers here.
```

---

6. Use the function `mean()` to calculate the average of the variable *female*. Please provide a full substantive interpretation of what this average means. (1 point)

```{r}
# Your coding answers here.
```

**Answer:** Change this line to type your answer here.

---

7. Use the function `mean()` to calculate the average of the variable *water*. Please provide a full substantive interpretation of what this average means. Make sure to provide the unit of measurement. (1 point)

```{r}
# Your coding answers here.
```

**Answer:** Change this line to type your answer here.

---

8. If we wanted to estimate the average causal effect of having a female politician on the number of new (and repaired) drinking water facilities: (1 point)
    a. What would be the treatment variable? Please provide the name of the variable
    b. What would be the outcome variable? Please provide the name of the variable

**Answer:** Change this line to type your answer here.

---

9. If we wanted to estimate the average causal effect of having a female politician on the number of new (and repaired) irrigation facilities: (1 point)
    a. What would be the treatment variable? Please provide the name of the variable
    b. What would be the outcome variable? Please provide the name of the variable

**Answer:** Change this line to type your answer here.

---

10. In both analyses above: (1 point)
    a. What would be the treatment group?
    b. What would be the control group?

**Answer:** Change this line to type your answer here.