---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 06 | Measuring Population Characteristics I}
fontsize: 12pt
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: ../../latex/style.tex
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- Github page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: [https://piazza.com/ucsd/winter2023/17221](https://piazza.com/ucsd/winter2023/17221)

## Before we start

**Recap:**

- We learned the definitions of Theory, Scientific Theory, and Hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, randomization, and ATE.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.8\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\\\
- Sample vs. Population\\
- Representative samples\\
- Random Sampling\\
- Random Treatment Assignment vs. \\
\,\,\,\, Random Sampling\\\\
- Exploring One Variable At a Time\\
\,\,\,\,\,\,\,- Table of frequencies\\
\,\,\,\,\,\,\,- Table of proportions\\
\,\,\,\,\,\,\,- Histogram\\
\,\,\,\,\,\,\,- Descriptive Statistics: mean, median, \\
\,\,\,\,\,\,\,\,\,\,\,standard deviation, and variance
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}


***
\vspace{.5cm}
\begin{mdframed}[linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{Why Do We Analyze Data?}\normalsize}\\\\
\color{gray}{1.} \color{black} MEASURE: \concept{To infer population}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\concept{characteristics via survey research}\\
\,\,\,\,\,\,\,\, - what proportion of constituents support\\
\,\,\,\,\,\,\,\,\,\,\,\, a particular policy?\\\\

\color{gray}{2.} \color{black}  PREDICT: \concept{To make predictions} \\
\,\,\,\,\,\,\,\, - who is the most likely candidate to win\\
\,\,\,\,\,\,\,\,\,\,\,\, an upcoming election?\\\\

\color{gray}{3.} \color{black}  EXPLAIN: \concept{To estimate the causal effect of}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\concept{a treatment on an outcome} \\
\,\,\,\,\,\,\,\, - what is the effect of small classrooms\\
\,\,\,\,\,\,\,\,\,\,\,\, on student performance?\\
\end{tabular}
\end{center}
\end{mdframed}

***
\vspace{.5cm}
\begin{mdframed}[linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{Why Do We Analyze Data?}\normalsize}\\\\
\color{gray}{1.} \color{black} MEASURE: \concept{To infer population}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\concept{characteristics via survey research}\\
\,\,\,\,\,\,\,\, - what proportion of constituents support\\
\,\,\,\,\,\,\,\,\,\,\,\, a particular policy?\\\\

\color{gray}{2.} \color{gray}  PREDICT: \textbf{To make predictions} \\
\,\,\,\,\,\,\,\, \color{gray}- who is the most likely candidate to win\\
\,\,\,\,\,\,\,\,\,\,\,\, \color{gray}an upcoming election?\\\\

\color{gray}{3.} \color{gray}  EXPLAIN: \textbf{To estimate the causal effect of}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\color{gray}\textbf{a treatment on an outcome} \\
\,\,\,\,\,\,\,\, \color{gray}- what is the effect of small classrooms\\
\,\,\,\,\,\,\,\,\,\,\,\, \color{gray}on student performance?\\
\end{tabular}
\end{center}
\end{mdframed}

## Sample vs. Population

\vspace{.2cm}

- We often want to know the characteristics of a large \concept{population}\ such as the residents of a country
\vspace{.2cm}

- Yet collecting data from every individual in the population is either prohibitively expensive or infeasible.
\vspace{.2cm}

- In the US, we try to collect data from each individual every ten years
    + The 2020 census cost \$14.2 billion, approximately (the population at that time was around 331 mi)
    + This is not feasible for research purposes!

- We use surveys to collect data from a small subset of observations in order to understand the population

## Sample vs. Population

- The subset of individuals chosen for study is called a \concept{sample}
\begin{center} \includegraphics[width=.6\textwidth]{../../images/sample.png}\end{center}
\vspace{.2cm}

- Researchers typically survey about 1,000 people to infer the characteristics of more than 200 million US citizens

- n=1,000, N=200 million
\vspace{-1cm}

## Representative Samples
- In survey research, the sample needs to be representative of the population of interest
\vspace{.2cm}

- \concept{Representative sample}\: Accurately reflects the characteristics of the population from which it is drawn
\vspace{.2cm}

- If the sample is not representative, our inferences regarding the population characteristics will be wrong
\vspace{-1cm}

## Representative Samples
- Are you a representative sample of US residents?
\vspace{.2cm}

- Are you a representative sample of UCSD students?
\vspace{.2cm}

- Are you a representative sample of UCSD Poli majors?
\vspace{.2cm}

- Are you a representative sample of POLI 30 D students?
\vspace{.2cm}

- What would be the best way to draw a representative sample of UCSD students?

## Representative Samples

- You should be careful about how representative your sample is:

\begin{center} \includegraphics[width=.6\textwidth]{../../images/truman.jpg}\end{center}

## Random Sampling

- The best way to draw a representative sample is to select individuals at *random* from the population.
\vspace{.4cm}\pause

\begin{mdframed}[backgroundcolor=white,linecolor=concept, linewidth=0.05cm]
\concept{Random sampling}\ makes the \textbf{sample} and the \textbf{target population} \textit{on average} identical.
\end{mdframed}
\vspace{.2cm}

- Random sampling: enables us to infer *valid* population characteristics from the sample,
\vspace{.2cm}

## Random Treatment Assignment vs. Random Sampling

- They both use a random process but are different concepts.
\vspace{.1cm}

- \concept{Random treatment assignment}\ means that treatment is assigned at random:
    + makes treatment and control groups comparable.

\vspace{.1cm}

- \concept{Random sampling}\ means that individuals are selected at random from the population into the sample:
    + makes the sample representative of the population.

\vspace{.1cm}

- For this class, we assume we are always studying a representative sample.

## Exploring One Variable At a Time

- Suppose we have collected data from a sample. Now what?

- To understand the content and distribution of each variable, we can:
    + Create a table of frequencies
    + Create a table of proportions
    + Create a histogram
    + Compute descriptive statistics
\vspace{.2cm}

- Let us return to the voting experiment
    +  data collected from a sample of registered voters in the state of Michigan
\vspace{-1cm}

## The *voting* dataset

\vspace{1cm}
\begin{table}
\begin{center}
\begin{tabular}{ll}											
\multicolumn{2}{l}{Unit of observation: registered voters}\\\\
\multicolumn{2}{l}{Description of variables:}\\\\
\midrule
variable & description\\
\midrule \addlinespace[1.5pt]
\textit{birth} &  year of birth of registered voter \\\addlinespace[1.5pt]
\textit{message} & whether registered voter received message:\\
& "yes", "no" \\\addlinespace[1.5pt]
\textit{voted} & whether registered voter voted:\\
& 1{=}voted, 0{=}didn't vote \\\addlinespace[1.5pt]
\midrule
\end{tabular}
\end{center}
\end{table}

## Table of Frequencies
- The \concept{frequency table}\ shows the values the variable takes and the number of times each value appears in the variable
\vspace{.2cm}

- R function: \fu{table()}

```{r, collapse=TRUE, comment='##', echo=FALSE} 
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
```
```{r, collapse=TRUE, comment='##'} 
table(voting$voted)
```

- Interpretation?
  \vspace{-1cm}

## Table of Proportions

- The \concept{table of proportions}\ shows the proportion of observations that take each value in the variable
\vspace{.2cm}

- The proportions in the table should add up to 1
\vspace{.2cm}

- R function: \fu{prop.table(table( ))}

```{r, collapse=TRUE, comment='##'} 
prop.table(table(voting$voted))
```

- Interpretation?
  \vspace{-1cm}

## Histogram

- The \concept{histogram}\ is the visual representation of a variable\'s distribution through bins of different heights
\vspace{.2cm}

- The position of the bins along the x-axis indicates the interval of values
\vspace{.2cm}

- The height of the bins indicates the frequency (or count) of the interval of values
\vspace{.2cm}

- R functions: \fu{hist()} or \fu{ggplot() + geom\_histogram()}

- Great for `quantitative` variables (the numeric R data types)

## Histogram

```{r, fig.height=2, fig.width=4, fig.align='center', collapse = T}
ggplot(voting, aes(x = birth)) + 
  geom_histogram(binwidth = 5, color = 'black') + theme_light()
```

- Interpretation?

\vspace{-2cm}

## Barplots

- The \concept{barplot}\ is similar to a histogram, but discretizes the variation
\vspace{.2cm}

- The position of the bins along the x-axis indicates a value
\vspace{.2cm}

- The height of the bins indicates the frequency (or count) of the values
\vspace{.2cm}

- R functions: \fu{barplot(table())} or \fu{ggplot() + geom\_bar()}

- Great for `qualitative` variables (numeric binary or character)

## Barplots

```{r, fig.height=2, fig.width=4, fig.align='center', collapse = T}
ggplot(voting, aes(x = voted)) + 
  geom_bar(aes(x = as.character(voted))) + theme_light()
```

- Interpretation?

\vspace{-2cm}

## Descriptive Statistics
  
- The \concept{descriptive statistics}\ of a variable numerically summarizes the main characteristics of its distribution
\vspace{.5cm}
- Measures of centrality \newline (center of the distribution): 
  + mean
  + median
\vspace{.2cm}
- Measures of spread \newline (amount of variation from the center): 
  + standard deviation 
  + variance
\vspace{-1cm}

## Mean

- The \concept{mean}\ of a variable equals the sum of the values across all observations divided by the total number of observations
\vspace{.2cm}

- What is the function in R?
  
- Example:
```{r, collapse=TRUE, comment='##'}
mean(voting$birth)
mean(voting$voted)
```

- Interpretations?
\vspace{-1cm}

## Median
- The \concept{median}\ of a variable is the value at the midpoint of the distribution that divides the data into two equal-size groups
\vspace{.2cm}

- When the variable contains an odd number of observations, the median is the middle value of the distribution 
\vspace{.2cm}

- When the variable contains an even number of observations, the median is the average of the two middle values
\vspace{-1cm}

## Median

- Example, if \textit{X}=\{10, 4, 6, 8, 22\}, what is the median of $X$?
  + First, we need to sort the values of $X$ in ascending order (as they would be in the distribution): \newline  $\{4, 6, 8, 10, 22\}$
  + The value in the middle of the distribution is $8$ so the median is $8$.

\vspace{.1cm}

- R function: \fu{median()}
```{r, collapse=TRUE, comment='##'}
median(voting$birth)
```

- Interpretations?

## Standard Deviation
- The \concept{standard deviation}\ of a variable is a measure of the spread of its distribution
$$ \text{sd(X)} = \sqrt{\frac{\sum^n_{i=1}(X_i-\overline{X})^2}{n}} $$
  
\begin{center}
\begin{tabular}{l}
- $sd(X)$ stands for the standard deviation of $X$\\
- $X_i$ is a particular observation of $X$\\
- $\overline{X}$ stands for the mean of $X$\\
- $n$ is the total number of observations in the variable\\
- $\sum^{n}_{i=1} (X_i{-}\overline{X})^2$ means the sum of all $(X_i{-}\overline{X})^2$\\ 
\,\,\,\,from $i = 1$ to $i = n$\\
\end{tabular}
\end{center}
\vspace{-1cm}

## Standard Deviation 

\begin{mdframed}[backgroundcolor=white,linecolor=concept, linewidth=0.05cm]
\begin{center} 
The \concept{standard deviation}\ of variable measures the average distance of the observations to the mean.
\end{center}
\end{mdframed}
\vspace{.2cm}

- The larger the standard deviation, the `flatter` the distribution

- It gives us a sense of the range of the data, especially when dealing with bell-shaped distributions

- In bell-shaped (normal) distributions, 95\% of the observations fall within two standard deviations from the mean

## Standard Deviation 

- R function: \fu{sd()}

```{r, fig.width=6, fig.height=3, collapse=TRUE, comment='##'}
sd(voting$birth)
```

- If \textit{birth} were normally distributed, about 95\% of the registered voters would have been born between 1927 and 1985:
  + $\overline{X}$ - 2 ${\times}$ sd(X) = 1956 - 2 $\times$ 14.5 = 1927 
  + $\overline{X}$ + 2 $\times$ sd(X) = 1956 + 2 $\times$ 14.5 = 1985

## Variance

- Another measure of the spread of the distribution

- The \concept{variance}\ of a variable is simply the square of the standard deviation

$$\text{var(X)} = [\text{sd(X)}]^2$$ 
  
\begin{center}
\begin{tabular}{l}
- $var(X)$ stands for the variance of $X$ \\
- $sd(X)$ stands for the standard deviation of $X$\\
\end{tabular}
\end{center}

## Variance

- R function: \fu{var()}

```{r, collapse=TRUE, comment='##'}
var(voting$birth)
```

-  Alternatively: \fu{sd()\^{}2}

```{r, fig.width=6, fig.height=3, collapse=TRUE, comment='##'}
sd(voting$birth)^2
```

-  We are usually better off using standard deviations as our measure of spread:
  + Same unit of measurement as the variable


## Summary

\vspace{0.2cm}

- **Today's Class:**
  + Sample vs. Population
  + Representative Samples and Random Sampling
  + Exploring a single variable

\vspace{0.2cm}

- **Next class:**
  + Correlations
  + Scatter-plots

\vspace{-2cm}

# Questions?

# See you in the next class!