---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 08 | Prediction I}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- GitHub page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: Not sure what the link is. Ask your TA!

- My mailbox disaster is not over, but things are in better shape now! Please let me know if I missed your email.

- \alert{PS 1 due today}!

- Note to self: Turn on the mic!

## Before we start

**Recap:** We learned:

- The definitions of theory, scientific theory, and hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.
- Sampling, descriptive statistics, and descriptive plots for one variable.
- Correlation between two continuous variables.

**Great job!**

- Do you have any questions about these contents?

***
\vspace{.5cm}
\begin{mdframed}[linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{Why Do We Analyze Data?}\normalsize}\\\\
\color{gray}{1.} \color{gray} MEASURE: \textbf{To infer population}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\color{gray} \textbf{characteristics via survey research}\\
\,\,\,\,\,\,\,\, \color{gray} - what proportion of constituents support\\
\,\,\,\,\,\,\,\,\,\,\,\, \color{gray} a particular policy?\\\\

\color{gray}{2.} \color{black}  PREDICT: \concept{To make predictions} \\
\,\,\,\,\,\,\,\, - who is the most likely candidate to win\\
\,\,\,\,\,\,\,\,\,\,\,\, an upcoming election?\\\\

\color{gray}{3.} \color{gray}  EXPLAIN: \textbf{To estimate the causal effect of}\\
\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\color{gray}\textbf{a treatment on an outcome} \\
\,\,\,\,\,\,\,\, \color{gray}- what is the effect of small classrooms\\
\,\,\,\,\,\,\,\,\,\,\,\, \color{gray}on student performance?\\
\end{tabular}
\end{center}
\end{mdframed}

***
\begin{center}
\begin{minipage}{0.85\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\\\
- Prediction and Linear Regression\\
- Example with Non-binary Target Variable:\\
\,\,\,\, Use income to predict education expenditure\\[.3cm]
\,\,\,\, 1. Load and explore data\\
\,\,\,\, 2. Identify X and Y\\
\,\,\,\, 3. What is the relationship between X and Y?\\
\,\,\,\,\,\,\,\,\,\,\,\,         - Create scatter plot \\
\,\,\,\,\,\,\,\,\,\,\,\,         - Calculate correlation\\
\,\,\,\, 4. Fit a linear model using the least squares method\\
\,\,\,\, 5. Interpret coefficients\\
\,\,\,\, 6. Make predictions\\
\,\,\,\, \color{gray} 7. Measure how well the model fits the data\color{black}
\end{tabular}
\end{center}
\vspace{.2cm}
\end{mdframed}
\end{minipage}
\end{center}

***
1. \concept{When estimating causal effects}
    + $X$ is the **treatment** variable (independent variable)
    \vspace{-.05cm}
    + $Y$ is the **outcome** variable (dependent variable)
    \vspace{-.05cm}
    + Aim: to estimate the effect of $X$ on $Y$
    \vspace{-.05cm}
    + Assumption: Treatment and control groups are comparable
    \vspace{-.05cm}
    + Best way of satisfying assumption: random treatment assignment

2. \pause \concept{When inferring population characteristics}
    + Aim: To infer the characteristics of $X$ in the population
    \vspace{-.05cm}
    + Assumption: sample is representative of the population
    \vspace{-.05cm}
    + Best way of satisfying assumption: Random sampling
\vspace{-1cm}

***
3. \concept{When making predictions}
    + When we need to use what we know to learn what we do not know.
    + $X$ is a variable(s) that we use as predictor(s) (independent variable[s]; also k.a. **features**)
    + $Y$ is our **target variable**: what we want to predict
    + $Y_i \ = \ f(X_i) + \varepsilon$; Where $i$ is a given observation of interest; $f(.)$ is the *shape* of the relationship; and $\varepsilon$ is the (inherent) error that the process entails.
    + Aim: to predict $Y$ as accurately as possible
    + Assumption: The shape of $f$. We will assume linear: $f(X) = \beta_0 + \beta_1 X_i$.
    + Best way to achieve our aim: To make $\textrm{R}^2$ as high as possible.
\vspace{-1cm}

## Using Income to Predict Education Expenditures in US States

- Today, we will analyze data on 1970 U.S. State Public-School Expenditures.

- Our goal is to model the relationship between per-capita income and per-capita education expenditures.

| **Variable** | **Meaning**                                 |
|--------------|---------------------------------------------|
| education    | Per-capita education expenditures, dollars. |
| income       | Per-capita income, dollars.                 |
| young        | Proportion under 18, per 1000.              |
| urban        | Proportion urban, per 1000.                 |
| states       | US State                                    |

\vspace{-1cm}

## Step 1: Load and Explore Data
```{r, collapse=TRUE, comment='##'}
educexp <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/educexp.csv")
head(educexp, 3)
```

- What is the unit of observation?
 
- For each variable: type and unit of measurement?

- Substantively interpret the first observation

\vspace{-1cm}
    
## Step 2: Identify the Dependent and Independent Variables

- The \concept{predictor (X)}\ is the variable we want to use to predict the outcome (Y).

- The \concept{target (Y)}\ is the variable that we want to predict.

- What are they?

```{r, fig.width=6, fig.height=2, echo=FALSE}
p1 <- ggplot(data = educexp, aes(x = income)) + 
  geom_histogram(bins = 7, color = 'black', fill = 'lightblue', alpha = 0.4) + labs(title = 'Per-Capita Income', y = '') + 
  theme_minimal()
p2 <- ggplot(data = educexp, aes(x = education)) + 
  geom_histogram(bins = 7, color = 'black', fill = 'lightblue', alpha = 0.4) + labs(title = 'Per-Capita Education Exp.', y = '') + 
  theme_minimal()
grid.arrange(p1, p2, nrow = 1)
```

## Step 3: What is the relationship between X and Y?

- Create \concept{scatter plot}\ to visualize the relationship between per-capita *income* and *education* expenditures.

```{r, fig.width=6, fig.height=3, echo=FALSE}
p <- ggplot(data = educexp, aes(x = income, y = education)) + 
#  geom_point(fill = 'lightblue', alpha = 0.4) + 
  geom_text(aes(label=states), size=2) + 
  labs(title = '', y = 'Per-Capita Education Expenditure', x = 'Per-Capita Income') + 
  theme_minimal()
p
```

\vspace{-2cm}

## Step 3: What is the relationship between X and Y?

```{r, fig.width=6, fig.height=2.5, echo=FALSE}
p
```

- The *Y variable* always goes in the *y-axis* and the *X variable* always goes in the *x-axis*.
\vspace{-0.1cm}
- Does the relationship look positive or negative?
\vspace{-0.1cm}
- Does the relationship look weekly or strongly linear?

\vspace{-2cm}

## Step 3: What is the relationship between X and Y?

- Let us now check the \concept{correlation}\ coefficient.

- It measures the direction and strength of the linear association between *income* and *education*.

\vspace{0.2cm}

```{r, collapse=TRUE, comment='##', myfntsize = TRUE}
cor(educexp$income, educexp$education)
```

- We find a moderately strong positive correlation

- Are we surprised by this number? Think about what we have seen in the scatter plot.

\vspace{-2cm}

## Step 3: What is the relationship between X and Y?

We learned so far:

- That an increase in per-capita *income* is associated with an increase in *education* expenditure.

\vspace{.2cm}

- What we want to know is: When *income* increases, then **by how much** the *education* expenditure is predicted to increase?

\vspace{.2cm}

- In general we care about: When $X$ increases by one unit, by how much is $Y$ predicted to change?

\vspace{.2cm}

- To answer this question, we will fit a `regression` line to summarize the relationship between $X$ and $Y$

\vspace{-1cm}

## Step 4: Fit a linear model using the least squares method

\vspace{-.1cm}

- Which line better summarizes the relationship?

\vspace{-.1cm}

```{r, fig.width=6, fig.height=2.5, echo=FALSE}
p1 <- ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
#  geom_text(aes(label=states), size=2) + 
  labs(title = '', y = '', x = '') + 
  geom_hline(yintercept = mean(educexp$education), color = 'blue', lwd = 1) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30))
p2 <- ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
#  geom_text(aes(label=states), size=2) + 
  labs(title = '', y = '', x = '') + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 30)) + 
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1)
grid.arrange(p1, p2, nrow = 1)
```

\vspace{-.2cm}

- The goal is the choose the line that best fits the data.
    + Which one do you think does that?

\vspace{-2cm}
    
## Step 4: Fit a linear model using the least squares method

- To choose the line best fits the data, we use \concept{the least squares method}.

- In \color{red}red\color{black}, you can see the *error* we make by approximating the *education* using the \color{blue}blue\color{black}\ trendline.

\vspace{-0.5cm}

```{r, fig.width=6, fig.height=2.5, echo=FALSE}
pred1dist <- rep(mean(educexp$education), dim(educexp)[1])
pred2dist <- as.numeric(lm('education ~ income', data = educexp)$fitted.values)
p1 <- ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
  geom_segment(aes(xend = income, yend = pred1dist), color = "red", lty = 3) +
#  geom_text(aes(label=states), size=2) + 
  labs(title = '', y = '', x = '') + 
  geom_hline(yintercept = mean(educexp$education), color = 'blue', lwd = 1) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30))
p2 <- ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
  geom_segment(aes(xend = income, yend = pred2dist), color = "red", lty = 3) +
#  geom_text(aes(label=states), size=2) + 
  labs(title = '', y = '', x = '') + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 30)) + 
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1)
grid.arrange(p1, p2, nrow = 1)
rm(pred1dist, pred2dist)
```

\vspace{-0.5cm}

- Which plot do you think is doing better?

\vspace{-1cm}

## Step 4: Fit a linear model using the least squares method

- We need to think about what *better* means: 
    + In the case of least square error, let the error in the prediction for a given US State $i$ be:

\vspace{-0.5cm}

$$ e_i \ = \ Y_i - \beta_0 - \beta_1 X_i $$

\vspace{-0.2cm}

- We need to find $\beta_0$ and $\beta_1$ that minimizes the sum of the squared error:

\vspace{-0.8cm}

$$ \min_{(\beta_0, \beta_1)} \sum_{i=1}^n e_i^2 \quad \text{which is the same as} \quad  \min_{(\beta_0, \beta_1)} \sum_{i=1}^n (Y_i - \beta_0 - \beta_1 X_i)^2 $$
\vspace{-0.5cm}

- The meaning of *least square method* should now be clear to you.

\vspace{-2cm}

## Step 4: Fit a linear model using the least squares method

- The fitted line is \concept{$\widehat{Y}$ = $\widehat{\beta}_0$ + $\widehat{\beta}_1$ X}
    + \concept{$\widehat{\beta}_0$}\ is the intercept
    + \concept{$\widehat{\beta}_1$}\ is the slope

\vspace{.2cm}

- If you learned that a line was $Y = mX + b$
    + think that $m$ is now $\widehat{\beta}_1$
    + think that $b$ is now $\widehat{\beta}_0$
\vspace{.2cm}

- **^** (called 'hat') stands for predicted or estimated
    + $\widehat{Y}$ is the predicted target outcome
    + $\widehat{\beta}_0$ and $\widehat{\beta}_1$ are the estimated coefficients

\vspace{-1cm}

## Step 4: Fit a linear model using the least squares method

- The R function to fit a linear model is the \fu{lm()}:

```{r, collapse=TRUE, comment='##', myfntsize = TRUE}
lm('education ~ income', data = educexp)
```

```{r, echo = FALSE, results='hide'}
mod <- lm(education ~ income, data = educexp)
```

- $\widehat{\beta}_0$ = `r round(mod$coef[1], digits=2)` and $\widehat{\beta}_1$ = `r round(mod$coef[2], digits=2)`

- The fitted line is $\widehat{Y}$ = `r round(mod$coef[1], digits=2)` + `r round(mod$coef[2], digits=2)` $X$ 

- More specifically: $\widehat{\textrm{education}}$ = `r round(mod$coef[1], digits=2)` + `r round(mod$coef[2], digits=2)` $\textrm{income}$

\vspace{-2cm}

## Step 4: Fit a linear model using the least squares method

- And to add a fitting line to a scatter plot, you can use \fu{abline()}\ or \fu{geom\_smooth()}. 
  + You are going to learn this in Labs 06 and 07.

```{r, fig.width=6, fig.height=2.2, mytiny = TRUE}
ggplot(data = educexp, aes(x = income, y = education)) + geom_text(aes(label=states), size=2) + 
  labs(title = '', y = 'Education Expenditure', x = 'Income') +
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1) + theme_minimal()
```

\vspace{-2cm}

## Step 5: Interpretation of Coefficients

- \concept{The intercept ($\widehat{\beta}_0$) is the $\widehat{Y}$ when $X$=0}.

```{r, fig.width=6, fig.height=2.5, echo=FALSE}
fit <- mod
ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_text(aes(label=states), size=2) + xlim(0, max(educexp$income)) + ylim(0, max(educexp$education)) +
  labs(title = '', y = 'Education Expenditure', x = 'Income') +
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1, fullrange = T) +
  theme_minimal() + 
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = mod$coef[1]), color = 'red')
```

\vspace{-.1cm}
   \vspace{-.2cm}
    + here: $\widehat{\beta}_0$ = `r round(mod$coeff[1], digits=2)`
    \vspace{-.1cm}


## Mathematical definition of $\widehat{\beta}_0$

\begin{align*}
\widehat{Y} &= \widehat{\beta}_0 + \widehat{\beta}_1\,\, X  & \color{gray}(\textrm{by definition})\\
\widehat{Y} &= \widehat{\beta}_0 + \widehat{\beta}_1\times 0 &\color{gray}(\textrm{if } X=0)\\
\widehat{Y} &= \widehat{\beta}_0 + 0 &\color{gray}(\textrm{if } X=0)\\
\widehat{Y} &= \widehat{\beta}_0 &\color{gray}(\textrm{if } X=0)
\end{align*}

\pause
> \vspace{.2cm} \begin{center} $\widehat{\beta}_0$ is the value of $\widehat{Y}$ when $X$=0 \end{center}


## Substantive interpretation of $\widehat{\beta}_0$

- Substitute X, Y, and $\widehat{\beta}_0$: 
  + $\widehat{\beta}_0$ = `r round(mod$coef[1], digits=2)` is the $\widehat{\textrm{\textit{education}}}$ when $\text{income} = 0$
  + When a State has 0 per-capita income, we predict that the per-capita expenditure in education will be `r round(mod$coef[1], digits=2)` dollars, on average
    + Sometimes, it is nonsensical (due to extrapolation)

\vspace{0.2cm}

- **Unit of measurement of $\widehat{\beta}_0$**? 
  + Same as $\overline{Y}$
  + In this case: *Y* is non-binary and measured in points so $\overline{Y}$ is measured in points and so is $\widehat{\beta}_0$

## Step 5: Interpretation of Coefficients



- Pick two points on the line, measure $\triangle \widehat{Y}$ and $\triangle X$ associated with the two points, calculate $\triangle \widehat{Y}$/$\triangle X$
  + Here: $\widehat{\beta}_1$ = $\textrm{0.06}$

\vspace{-1cm}

## Step 5: Interpretation of Coefficients

- \concept{The slope ($\widehat{\beta}_1$) is the $\triangle \widehat{Y}$ associated with $\triangle X$=1}

\begin{center}\includegraphics[height=.4\textheight]{../../images/change.png}\end{center}

\vspace{-.5cm}

\begin{eqnarray*}
\triangle \widehat{Y} &=& \widehat{Y}_{\textrm{final}} {-} \widehat{Y}_{\textrm{initial}}\\
\triangle X &=& X_{\textrm{final}} {-} X_{\textrm{initial}}\\
\end{eqnarray*}

\vspace{-2cm}

## Mathematical definition of $\widehat{\beta}_1$

\small

\vspace{.2cm}
\begin{align*}
\triangle \widehat{Y} &= \widehat{Y}_{\textrm{final}} {-} \widehat{Y}_{\textrm{initial}} & \color{gray}(\textrm{by definition}) \\
\triangle \widehat{Y} &= (\widehat{\beta}_0 + \widehat{\beta}_1 X_{\textrm{final}}) - (\widehat{\beta}_0 + \widehat{\beta}_1 X_{\textrm{initial}}) & \color{gray}(\textrm{since } \widehat{Y}_{\textrm{final}}= \widehat{\beta}_0 + \widehat{\beta}_1 X_{\textrm{final}} \,\,\,\,\\
&& \color{gray}\textrm{and } \widehat{Y}_{\textrm{initial}}= \widehat{\beta}_0 + \widehat{\beta}_1 X_{\textrm{initial}}) \\
\triangle \widehat{Y} &= \widehat{\beta}_0 - \widehat{\beta}_0 + \widehat{\beta}_1 \,(X_{\textrm{final}}{-} X_{\textrm{initial}}) & \color{gray}(\textrm{rearranging terms}) \\
\triangle \widehat{Y} &= \widehat{\beta}_1 \,(X_{\textrm{final}}{-} X_{\textrm{initial}}) & \color{gray}(\textrm{since }\widehat{\beta}_0 - \widehat{\beta}_0 = 0)\\
\triangle \widehat{Y} &= \widehat{\beta}_1 \,(\triangle X) & \color{gray}(\textrm{since } \triangle X = X_{\textrm{final}} - X_{\textrm{initial}}) \\
\triangle \widehat{Y} &= \widehat{\beta}_1 \times 1 &\color{gray}(\textrm{if } \triangle X=1)\\
\triangle \widehat{Y} &= \widehat{\beta}_1 &\color{gray}(\textrm{if } \triangle X=1)
\end{align*}

\normalsize

\vspace{-.1cm}\pause
> \vspace{.2cm} $\widehat{\beta}_1$ is the value of $\triangle \widehat{Y}$ associated with $\triangle X$ = 1

## Substantive interpretation of $\widehat{\beta}_1$

- Start with the mathematical definition: 
  + $\widehat{\beta}_1$ is the $\triangle \widehat{Y}$ associated with $\triangle X$=1

- Substitute $X$, $Y$, and $\widehat{\beta}_1$: 
  + $\widehat{\beta}_1$ = `r round(mod$coef[2], digits=2)` is the $\triangle \widehat{\textrm{\textit{education}}}$ associated with $\triangle$*income*=1
    + An increase in income of 1 dollar is associated with a predicted increase in education expenditures of 6 cents, on average

- **Unit of measurement of $\widehat{\beta}_1$**?
  + Same as $\triangle \overline{Y}$
    + In this case: *Y* is non-binary and measured in points so $\triangle \overline{Y}$ is measured in points and so is $\widehat{\beta}_1$

***
\begin{mdframed}[linecolor=titleline, linewidth=0.05cm, innerleftmargin=.5cm,innerrightmargin=.5cm]
\vspace{.3cm} 
\begin{center}\color{concept}\tip{The fitted line is}:\color{text}\end{center}
$$\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$$

\begin{tabular}{l}
- $\widehat{\beta}_0$ (beta-zero-hat) is the estimated intercept coefficient\\
\,\,\,\,\,\,\,\,\,\,\,the $\widehat{Y}$ when $X{=}\textrm{0}$\\
\,\,\,\,\,\,\,\,\,\,\,(in same unit of measurement as $\overline Y$)\\\\

- $\widehat{\beta}_1$ (beta-one-hat) is the estimated slope coefficient\\
\,\,\,\,\,\,\,\,\,\,\,the $\triangle \widehat{Y}$ associated with $\triangle X{=}\textrm{1}$\\
\,\,\,\,\,\,\,\,\,\,\,(in the same unit of measurement as $\triangle\overline Y$)
\end{tabular}
\vspace{.2cm}
\end{mdframed}

## Step 6: Make Predictions

- Now that we have found the line that best summarizes the relationship between X and Y, we can use it to make predictions
\vspace{.2cm}

- There are two types of predictions that we might be interested in:
\vspace{.3cm}

1. predict $\widehat{Y}$ based on $X$: \concept{$\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$}

\vspace{0.3cm}

2. predict $\triangle \widehat{Y}$ associated with $\triangle X$: \concept{$\triangle \widehat{Y} = \widehat{\beta}_1 \triangle X$}

\vspace{-1cm}

## Step 6: Make Predictions

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\noindent To predict $\widehat{Y}$ based on $X$: \color{titleline} $\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$
\end{mdframed}

- \underline{Example 1}: Suppose we are back in the 70s. Imagine you lived in a State where the per-capita income is \$ 3,500. What would the education expenditure be? 

\vspace{-.9cm}
\begin{eqnarray*}
\widehat{\textrm{education}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \,\,\textrm{\,\,income} \\
\widehat{\textrm{education}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \times \textrm{3,500} \color{gray} \textrm{\,\, (if income = 3,500)} \\ 
\widehat{\textrm{education}} &=& \textrm{`r round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*3500`} \\
\end{eqnarray*}
\vspace{-1.5cm}\pause

- Answer: If the income per-capita was \$ 3,500, the the education expenditure would be \$ `r round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*3500`, on average.

\vspace{-1cm}    

## Step 6: Make Predictions

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm]
\noindent To predict $\triangle \widehat{Y}$ associated with $\triangle X$: \color{titleline} $\triangle \widehat{Y} = \widehat{\beta}_1 \, \triangle \text{X}$
\end{mdframed}
- \underline{Example 2}: Suppose the per-capita income rises by \$100. By how much would we predict that the education expenditure would change?

\vspace{-.9cm}
\begin{eqnarray*}
\triangle\widehat{\textrm{education}} &=& \textrm{`r round(fit$coef[2], digits=2)`} \,\,\triangle\textrm{income} \\
\triangle \widehat{\textrm{education}} &=& \textrm{`r round(fit$coef[2], digits=2)`} \times \textrm{100} \color{gray} \textrm{\,\, (if $\triangle$income = 100)} \color{black} \\
\triangle \widehat{\textrm{education}} &=& \textrm{`r round(fit$coef[2], digits=2)*100`} \\
\end{eqnarray*}
\vspace{-1.6cm}

- Answer: An increase of \$100 in per-capita income is associated with a predicted increase of \$6.00 in the average education expenditure

\vspace{-1cm}    

## Summary

- **Today's Class:**
  + How to summarize the relationship between X and Y with a line: \fu{lm()}\ and \fu{geom\_smooth()}.
  + How to interpret the two estimated coefficients: ($\widehat{\beta}_0$ and $\widehat{\beta}_1$) when outcome variable is non-binary.
  + How to make predictions with the fitted line: 
    + Predict $\widehat{Y}$ based on $X$ and predict.
    + Predict $\triangle\widehat{Y}$ based on $\triangle X$

\vspace{0.2cm}

- **Next class**:
  + Another example of how to use the linear model to make predictions, but with binary outcomes.
  + How to measure how well the model fits the data with $\textrm{R}^2$.
  
\vspace{-1cm}


# Questions?

# See you in the next class!