---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 06 | R Plots and R Data Analysis II}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: '../../latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, echo=FALSE, results='hide', include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
library(tidyverse)
```

## Before we start

**Announcements:**

- GitHub page: [**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: The link in the slides needs to be fixed. Check with instructors for an alternative link.

## Before we start

**Recap:** In the Lab sessions, you learned:

- How to install R and R Studio on your computer.
- How to do basic math operations in R.
- How to do basic vector and data.frame operations in R.
- How to install packages and work with R Markdown.
- How to work with advanced R objects and create histograms.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 05}}}\normalsize\\\\
- Barplots\\
- Violinplots\\
- Scatterplots\\
- Correlation\\
- Bivariate Regression\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# Getting started

## Getting started

- To get started, we need to load the datasets we will need in the lab.

- We also need to load the `tidyverse` package, which has all the R functions we use.

- Let's do it, then!

## Getting started - tidyverse

- Loading the `tidyverse` library:

```{r}
library(tidyverse)
```

## Getting started - Education expenditure data

```{r}
educexp <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/educexp.csv")
head(educexp)
```

## Getting started - Chile survey data

```{r}
chilesurv <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/survchile.csv")
head(chilesurv)
```

## Getting started - Voting

```{r}
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
head(voting)
```

# Intro to plots (revisited)

## Intro to plots

- For plots, we will use a package called `ggplot2`.

- Here is a good [cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf). This is a great thing to print and has close by when creating plots.

- `ggplot2` is based on the `grammar of graphs`. But what is this?

## Intro to plots

- In the abstract, the grammar of graphs is a decomposition of plots in its main features.

- In essence, every plot has the following:

1. A dataset
2. A coordinated system
3. A geometric shape

- And different plots are different compositions of these three key ingredients.

# Barplots

## Barplots

- Barplots is great for representing a binary variable. The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       mapping = aes(x = variable_x_name)) +
  geom_bar()
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- You need to add `dataset` and the `variable_x_name`.

## Barplots

```{r, fig.height=2.5, fig.width=2.5, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = vote)) +
  geom_bar()
```

\vspace{-2cm}

## Barplots

```{r, fig.height=2.5, fig.width=2.5, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = vote)) +
  geom_bar(color = 'black', fill = 'lightblue', alpha = 0.6) +
  labs(title = 'Barplot of Vote Intention', x = 'Vote Intention', y = '') + theme_minimal()
```

\vspace{-2cm}

# Plots for two variables

## Plots for two variables

- Most fun things are when we plot one variable against another.

- This is because exploring one variable may be fun, but it could be more informative.

- We want to find relationships between variables!

## Plots for two variables

- For this case, whenever a variable is binary or non-binary, we have three combinations with respective plots:

1. Binary x Binary $\rightarrow$ Mosaic Plots
2. Binary x Non-binary $\rightarrow$ Violin Plots
3. Non-binary x non-binary $\rightarrow$ Scatter Plots

# Mosaic Plots

## Mosaic Plots

- Type of barplot for two non-binary variables. The syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       mapping = aes(x = treatment_var, 
                     fill = as.factor(outc_var))) + 
  geom_bar(position = 'fill')
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- And you need to change the `dataset`, the `outc_var`, and the `treatment_var`.

- The mosaic plots make the relationship between a binary treatment and a binary control very clear!

## Mosaic Plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = voting, aes(x = message, fill = as.factor(voted))) +
  geom_bar(position = 'fill')
```

\vspace{-2cm}

## Mosaic Plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = voting, aes(x = message, fill = as.factor(voted))) +
  geom_bar(position = 'fill', alpha = 0.7, color = 'gray') + 
  scale_fill_manual(values = c('red', 'blue'), name = 'Voted') + 
  labs(title = 'Mosaicplot of Vote x Message', 
       x = 'Message Status', y = 'Vote Intention') + theme_minimal()
```

\vspace{-2cm}

# Violin Plots

## Violin Plots

- Violin plot is excellent when you want to check how a non-binary variable and a binary variable are related. The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       mapping = aes(x = as.factor(binary_var), 
                     y = nonbin_var)) + 
  geom_violin()
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- And you need to change the `dataset`, the `binary_var`, and the `nonbin_var`.

- But they tend to look somewhat ugly...

\vspace{-1cm}

## Violin Plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = as.factor(vote), y = statusquo)) +
  geom_violin()
```

\vspace{-2cm}

## Violin Plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = as.factor(vote), y = statusquo)) +
  geom_violin(fill = 'pink', alpha = 0.7, bw = 0.6, draw_quantiles = 0.5) + 
  labs(title = 'Vote x Status-quo Satisfaction', 
       x = 'Vote', y = 'Status-quo') + theme_minimal()
```

\vspace{-2cm}

# Scatter Plots

## Scatter plots

- Scatter plots are great for two non-binary variables. The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       mapping = aes(x = indep_var, y = dep_var)) + 
  geom_point()
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- And you need to change the `dataset`, the `indep_var`, and the `dep_var`.

- They make the relationship between two non-binary variables very clear!

- And you can add a trend line.

\vspace{-1cm}

## Scatter plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = educexp, aes(x = income, y = education)) +
  geom_point()
```

\vspace{-2cm}

## Scatter plots

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
  #geom_text(aes(label=states), size=2) + 
  labs(title = '', y = 'Per-Capita Education Expenditure', x = 'Per-Capita Income') + 
  theme_minimal()
```

\vspace{-2cm}

## Scatter plots with trend line

- To add a trend line, you do the following:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       mapping = aes(x = indep_var, y = dep_var)) + 
  geom_point() + 
  geom_smooth(se = F, formula = 'y ~ x')
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- And you need to change the `dataset`, the `indep_var`, and the `dep_var`.

- It adds a non-linear trend line called `loess`. To change that, add the `method = 'lm'` parameter!

## Scatter plots with trend line

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = educexp, aes(x = income, y = education)) +
  geom_point() + geom_smooth(se = F, formula = 'y ~ x')
```

\vspace{-2cm}

## Scatter plots with trend line

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = educexp, aes(x = income, y = education)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
  labs(title = '', y = 'Education Expenditure', x = 'Income') +
  geom_smooth(formula = 'y ~ x', method = 'lm', 
              se = F, color = 'blue', lwd = 1) + theme_minimal()
```

\vspace{-2cm}

## Scatter plots with binary dependent

- If you want to do a scatterplot with a binary response variable, like the one in class, you need to add some `jitter`!

- The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
ggplot(data = dataset, 
       aes(x = indepvar, y = bindepvar)) + 
  geom_jitter(height = amount_x_jitter, 
              width = amount_y_jitter)
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- You need to change the `dataset`, the `indepvar`, and the `bindepvar`, the `amount_x_jitter`, and `amount_y_jitter` for a number between zero and one.

- It is great for a binary response variable!

## Scatter plots with binary dependent

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = statusquo, y = voteYES)) +
  geom_jitter(height = 0.4, width = 0)
```

\vspace{-2cm}

## Scatter plots with binary dependent

```{r, fig.height=2.5, fig.width=3, fig.align='center', mytiny = TRUE}
ggplot(data = chilesurv, aes(x = statusquo, y = voteYES)) + 
  geom_jitter(fill = 'lightblue', alpha = 0.5, height = 0.4, width = 0) + 
  labs(title = '', y = 'Intention to vote YES\nin the plebiscite', 
       x = 'Status-Quo Scale') + theme_minimal()
```

\vspace{-2cm}

# Correlation

## Correlation

- Computing correlations in R is very easy. The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
  cor(dataset$var1, dataset$var2)
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- You need to change the `dataset`, the `var1`, and the `var2`.

- By the way, the order of variables in the correlation does not matter.

## Correlation

- Correlation between education expenditure and income:

```{r, collapse=TRUE}
cor(educexp$education, educexp$income)
```

- **Your turn:** what is the correlation between education expenditure and the proportion of young people?

# Bivariate regression

## Bivariate regression

- Bivariate regression is to fit the function $Y = \beta_0 + \beta_1 X + \varepsilon$. The syntax is simple:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
  lm(Y ~ X, data = dataset)
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- You need to change the `dataset`, the `Y` variable, and the `X` variable.

- `Y ~ X` is called the formula for your regression.

- It spits out a pair of numbers for $\beta_0$ (we call $\widehat{\beta}_0$) and $\beta_1$ (we call $\widehat{\beta}_1$).

## Bivariate regression

- A bivariate regression for education expenditure explained by income:

```{r, collapse=TRUE}
lm(education ~ income, data = educexp)
```

- **Your turn:** what are the bivariate regression estimates for education expenditure and the proportion of young people?


***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Today's Lab}}}\normalsize\\
- Barplots\\
- Violinplots\\
- Scatterplots\\
- Correlation\\
- Bivariate Regression\\\\
\multicolumn{1}{c}{\large\concept{\textbf{Next Lab}}}\normalsize\\
- More plots and analysis\\
- Some data wrangling\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}
  
# Questions?

# See you in the next lab!