---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 07 | Measuring Population Characteristics II}
fontsize: 12pt
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- Github page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: [https://piazza.com/ucsd/winter2023/17221](https://piazza.com/ucsd/winter2023/17221)

- In a midst of a mailbox disaster now. Will check all your emails by Thursday evening!

- If you don't see me wearing a mic, tell me!

## Before we start

**Recap:** We learned:

- The definitions of theory, scientific theory, and hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.
- Sampling, descriptive statistics, and descriptive plots for one variable.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.8\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\\\
- Exploring the Relationship \\
\,\,\,\,Between Two Variables\\
\,\,\,\,\,\,\,- Scatter plots\\
\,\,\,\,\,\,\,- Correlations\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

## Scatter Plots

- A \concept{scatter plot}\ enables us to visualize the relationship between two variables by plotting one against the other
\vspace{.2cm}

```{r, cache=FALSE, fig.width=10, fig.height=5, echo=F, out.width = '80%', fig.align='center'}
star <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/STAR.csv")
par(mfrow = c(1,2),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular
plot(star$reading, star$math, col="#d4d4d4", pch=19, main="", axes=F, col.lab="gray17", bty="n", ylab="", xlab="")
mtext("Y", side = 2, line = 1, outer = FALSE, cex=1.5, col.lab="gray17", at=780, col="gray17")
mtext("X", side = 1, line = 1, outer = FALSE, cex=1.5, col.lab="gray17", adj=1, col="gray17")
box(col="gray17")

plot(star$reading, -star$math, col="#d4d4d4", pch=19, main="", axes=F, col.lab="gray17", bty="n", ylab="", xlab="")
mtext("Y", side = 2, line = 1, outer = FALSE, cex=1.5, col.lab="gray17", at=-509, col="gray17")
mtext("X", side = 1, line = 1, outer = FALSE, cex=1.5, col.lab="gray17", adj=1, col="gray17")
box(col="gray17")
```
\vspace{-1cm}

## Scatter Plots

\vspace{-.2cm}
Imagine we have two variables:
\begin{tabular}{ccl}
 X & Y & \\
 \cline{1-2}
 4 & 2 & \\
 8 & 5 & \\
 10 & 3 & \\
\end{tabular}\pause

We can create the scatter plot by plotting one point at a time.

## Scatter Plots

Imagine we have two variables:
\begin{tabular}{ccl}
 X & Y & \\
 \cline{1-2}
 4 & 2 & >> First, let's plot this point: \concept{(x$_1$, y$_1$) = (4,2)}\\
 8 & 5 & \\
 10 & 3 & \\
\end{tabular}

\vspace{-2cm}

## Scatter Plots

Imagine we have two variables:
\begin{tabular}{ccl}
 X & Y & \\
 \cline{1-2}
 4 & 2 & >> First, let's plot this point: \concept{(x$_1$, y$_1$) = (4,2)}\\
 8 & 5 & \\
 10 & 3 & \\
\end{tabular}

\vspace{-2cm}

## Scatter Plots

```{r, echo=FALSE, fig.width=8, fig.height=3.5}
Y <- c(2,5,3)
X <- c(4,8,10)
par(mfrow = c(1,1),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular

plot(X,Y, col="white", ylim=c(0,7), xlim=c(0,12), xaxt = "n", yaxt="n", yaxs="r", xaxs="r", ylab="", xlab="", axes=F, col.lab="gray17", bty="n")

mtext("Y", side = 2, line = 1, outer = FALSE, cex=.8, col.lab="gray17", at=7, col="gray17")
mtext("X", side = 1, line = 1, outer = FALSE, cex=.8, col.lab="gray17", adj=1, col="gray17")

axis(side = 1,at = c(0,1,2,3,4,5,6,7,8,9,10,11),c(0,1,2,3,4,5,6,7,8,9,10,11), col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
axis(side = 2,at = c(0,1,2,3,4,5,6),c(0,1,2,3,4,5,6),col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
box(col="gray17")

# first point
lines(x=c(X[1],X[1]), y=c(-0.3,Y[1]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[1]), y=c(Y[1],Y[1]), col="gray", lty=2, lwd=1)
points(X[1],Y[1], col="#ef3119", lwd=1, pch=19)
text(X[1],Y[1], "(4,2)", col="#ef3119", pos=4, offset=1, cex=1.3)
```
 
\vspace{-2cm}

## Scatter Plots

\vspace{-.2cm}
Imagine we have two variables:
\begin{tabular}{ccl}
 X & Y & \\
 \cline{1-2}
 4 & 2 & >> First, let's plot this point: \concept{(x$_1$, y$_1$) = (4,2)}\\
 8 & 5 & >> Now, let's plot this point: \concept{(x$_2$, y$_2$) = (8,5)}\\
 10 & 3 & \\
\end{tabular}

\vspace{-2cm}

## Scatter Plots

```{r, echo=FALSE, fig.width=8, fig.height=3.5}
Y <- c(2,5,3)
X <- c(4,8,10)
par(mfrow = c(1,1),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular

plot(X,Y, col="white", ylim=c(0,7), xlim=c(0,12), xaxt = "n", yaxt="n", yaxs="r", xaxs="r", ylab="", xlab="", axes=F, col.lab="gray17", bty="n")

mtext("Y", side = 2, line = 1, outer = FALSE, cex=.8, col.lab="gray17", at=7, col="gray17")
mtext("X", side = 1, line = 1, outer = FALSE, cex=.8, col.lab="gray17", adj=1, col="gray17")

axis(side = 1,at = c(0,1,2,3,4,5,6,7,8,9,10,11),c(0,1,2,3,4,5,6,7,8,9,10,11), col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
axis(side = 2,at = c(0,1,2,3,4,5,6),c(0,1,2,3,4,5,6),col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
box(col="gray17")

# first point
lines(x=c(X[1],X[1]), y=c(-0.3,Y[1]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[1]), y=c(Y[1],Y[1]), col="gray", lty=2, lwd=1)
points(X[1],Y[1], col="#ef3119", lwd=1, pch=19)
#text(X[1]+1.5,Y[1], "(4,2)", col="#ef3119")
text(X[1],Y[1], "(4,2)", col="#ef3119", pos=4, offset=1, cex=1.3)

# second point
lines(x=c(X[2],X[2]), y=c(-0.3,Y[2]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[2]), y=c(Y[2],Y[2]), col="gray", lty=2, lwd=1)
points(X[2],Y[2], col="#ef3119", lwd=1, pch=19)
#text(X[2]+.7,Y[2]+1, "(8,5)", col="#ef3119")
text(X[2],Y[2], "(8,5)", col="#ef3119", pos=4, offset=1, cex=1.3)
```

\vspace{-2cm}

## Scatter Plots

Imagine we have two variables:
\begin{tabular}{ccl}
 X & Y & \\
 \cline{1-2}
 4 & 2 & >> First, let's plot this point: \concept{(x$_1$, y$_1$) = (4,2)}\\
 8 & 5 & >> Now, let's plot this point: \concept{(x$_2$, y$_2$) = (8,5)}\\
 10 & 3 & >> Finally, let's plot this point: \concept{(x$_3$, y$_3$) = (10,3)}\\
\end{tabular}

\vspace{-2cm}

## Scatter Plots

```{r, echo=FALSE, fig.width=8, fig.height=3.5}
Y <- c(2,5,3)
X <- c(4,8,10)

par(mfrow = c(1,1),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular

plot(X,Y, col="white", ylim=c(0,7), xlim=c(0,12), xaxt = "n", yaxt="n", yaxs="r", xaxs="r", ylab="", xlab="", axes=F, col.lab="gray17", bty="n")

mtext("Y", side = 2, line = 1, outer = FALSE, cex=.8, col.lab="gray17", at=7, col="gray17")
mtext("X", side = 1, line = 1, outer = FALSE, cex=.8, col.lab="gray17", adj=1, col="gray17")

axis(side = 1,at = c(0,1,2,3,4,5,6,7,8,9,10,11),c(0,1,2,3,4,5,6,7,8,9,10,11), col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
axis(side = 2,at = c(0,1,2,3,4,5,6),c(0,1,2,3,4,5,6),col="white", col.ticks="gray", col.axis="gray17", tck=-0.02)
box(col="gray17")

# first point
lines(x=c(X[1],X[1]), y=c(-0.3,Y[1]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[1]), y=c(Y[1],Y[1]), col="gray", lty=2, lwd=1)
points(X[1],Y[1], col="#ef3119", lwd=1, pch=19)
text(X[1],Y[1], "(4,2)", col="#ef3119", pos=4, offset=1, cex=1.3)

# second point
lines(x=c(X[2],X[2]), y=c(-0.3,Y[2]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[2]), y=c(Y[2],Y[2]), col="gray", lty=2, lwd=1)
points(X[2],Y[2], col="#ef3119", lwd=1, pch=19)
text(X[2],Y[2], "(8,5)", col="#ef3119", pos=4, offset=1, cex=1.3)

# third point
lines(x=c(X[3],X[3]), y=c(-0.3,Y[3]), col="gray", lty=2, lwd=1)
lines(x=c(-0.3,X[3]), y=c(Y[3],Y[3]), col="gray", lty=2, lwd=1)
points(X[3],Y[3], col="#ef3119", lwd=1, pch=19)
text(X[3],Y[3], "(10,3)", col="#ef3119", pos=4, offset=1, cex=1.3)
```

\vspace{-2cm}

## Scatter Plots

- R functions: \fu{plot()} or \fu{ggplot() + geom\_point()}
\vspace{.2cm}

- How many arguments are required?
    + two; the two variables

## Scatter Plots

- Let us use the data from Project STAR:
    ```{r, collapse=TRUE, comment='##', echo=FALSE} 
    star <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/STAR.csv") 
    ```
    ```{r, collapse=TRUE, comment='##'} 
    head(star, 3) # shows first observations
    ```

- Unit of observation?
\vspace{-.1cm}
    + students; each observation represents a student

- Unit of measurement of *reading* and *math*?
\vspace{-.1cm}
    + points

## Scatter Plots

```{r, cache=FALSE,  fig.width=4, fig.height=4, echo=F, out.width = '50%', fig.align='center'}
ggplot(star, aes(x = reading, y = math)) + 
  geom_point() + theme_light()
```

- What can we learn from this scatter plot?

\vspace{-2cm}

## Correlation Coefficient

- The \concept{correlation coefficient} is a statistic that summarizes the relationship between two variables with a number
    + denoted as cor(X,Y)
\vspace{.2cm}

- cor(X,Y) summarizes the \concept{direction}\ and the \concept{strength}\ of the \concept{linear association}\ between X and Y

- cor(X,Y) ranges from -1 to 1

## Correlation Coefficient

The sign reflects the \concept{direction}\ of the linear association:

+ cor(X,Y) > 0 (tends to see one **in**creasing when the other increases)
+ cor(X,Y) < 0 (tends to see one **de**creasing when the other increases)
\vspace{.2cm}

\pause The absolute value reflects the \concept{strength}\ of the linear association:

+ |cor(X,Y)| = 0 if there is no linear association
+ |cor(X,Y)| = 1 if there is a perfect linear association
+ |cor(X,Y)| increases as the linear association becomes stronger

\vspace{-1cm}

## Correlation Coefficient

\concept{Direction}\ of the linear association between two variables:
\vspace{.5cm}

```{r, cache=FALSE, fig.width=10, fig.height=5, echo=F, out.width = '80%', fig.align='center'}
par(mfrow = c(1,2),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular
plot(star$reading, star$math, col="#d4d4d4", pch=19, main="", axes=F, col.lab="gray17", bty="n", ylab="", xlab="")
box(col="gray17")
plot(star$reading, -star$math, col="#d4d4d4", pch=19, main="", axes=F, col.lab="gray17", bty="n", ylab="", xlab="")
box(col="gray17")
```

\begin{center}
\begin{tabular}{rcl} 
positive linear association & vs. & negative linear association\\
positive correlation  & vs. & negative correlation\\
\end{tabular}
\end{center}
\vspace{-1cm}

## Correlation Coefficient

\concept{Strength}\ of the linear association between two variables:
\vspace{.5cm}

```{r, cache=FALSE, fig.width=10, fig.height=5, echo=F, out.width = '80%', fig.align='center'}
par(mfrow = c(1,2),
          oma = c(0,0,0,0) + 0.1, # outer margin: bottom, left, top, right
          mar = c(3,3,1,1) + 0.1, # inner margin, default: c(5,4,4,2) + 0.1
          mgp = c(2,0.5,0), # margin line for axis title, default: c(3,1,0)
    cex=1, adj=1, # 1: right-justified
    yaxs="r", xaxs="r", las=1) # axis choice: "i" internal, "r" regular

set.seed(123)
x <- rnorm(1000, mean=50, sd=3)
y <- x + rnorm(1000, mean=50, sd=22)
plot(x,y, col="#c7c7c7", ylab="", xlab="", cex=.6, xaxt = "n", yaxt="n", yaxs="r", xaxs="r", axes=F, col.lab="gray17", bty="n", pch=19)
box(col="gray17")

x <- rnorm(1000, mean=50, sd=10)
y <- x + rnorm(1000, mean=50, sd=7)
plot(x,y, col="#c7c7c7", ylab="", xlab="", cex=.6, xaxt = "n", yaxt="n", yaxs="r", xaxs="r", axes=F, col.lab="gray17", bty="n", pch=19)
box(col="gray17")
```

\vspace{.2cm}
\begin{center}
\begin{tabular}{rcl} 
\textbf{weak} linear association & vs. & \textbf{strong} linear association\\
absolute value closer to 0 & vs. & absolute value closer to 1\\
\end{tabular}
\end{center}
\vspace{-1cm}

## Correlation Coefficient

```{r, cache=FALSE,  fig.width=6, fig.height=6, echo=F, out.width = '50%', fig.align='center'}
ggplot(star, aes(x = reading, y = math)) +
  geom_point() + theme_light()
```

- Do you expect the correlation between *reading* and *math* grades to be positive or negative?

\vspace{-2cm}

## Correlation Coefficient

```{r, cache=FALSE,  fig.width=6, fig.height=6, echo=F, out.width = '50%', fig.align='center'}
ggplot(star, aes(x = reading, y = math)) +
  geom_point() + theme_light()
```

- Do you expect the absolute value of the correlation between *reading* and *math* to be closer to 1 or to 0?

\vspace{-2cm}

## Correlation Coefficient
- R function: \fu{cor()}
\vspace{.2cm}
- How many required arguments?
    + two; the two variables
\vspace{.2cm}

- Does the order of the arguments matter? 
    + no; cor(X,Y) = cor(Y,X)
\vspace{.2cm}

- What is the code to compute the correlation between *reading* and *math*?
    - Answer:
        ```{r, collapse=TRUE, comment='##'}   
        cor(star$reading, star$math) 
        ```
    + Is the correlation what we expected?

\vspace{-2cm}

## Correlation Coefficient
```{r, cache=FALSE,  fig.width=6, fig.height=2.9, echo=F, out.width = '80%', fig.align='center'}
par(mfrow = c(1,2),
    oma = c(1,0,1,0) + 0.1, # bottom, left, top, right # outer margins
    mar = c(0,1,0,1) + 0.1,   
    mgp = c(2,.5,0), cex=.45, adj=1, yaxs="r", xaxs="r",  yaxt="n", xaxt="n")

x <- rnorm(1000, mean=50, sd=10)
e <- rnorm(1000, mean=0, sd=7)
e2 <- rnorm(1000, mean=100, sd=40)
y1 <- x + e
y2 <- 5*x + e2

plot(x,y1, ylim=c(-100,200), xlim=c(0,100), ylab="", xlab="", col="#d4d4d4", pch=19, yaxt="n", xaxt="n", cex.lab=1.5, col.lab="gray17", bty="n")
box(col="gray17")
abline(lm(y1~x), lty="dashed", col="#ef3119", lwd=1.4)

plot(x,y2, ylab="", xlab="", ylim=c(0,600), xlim=c(0,100), col="#d4d4d4", pch=19, yaxt="n", xaxt="n", cex.lab=1.5, col.lab="gray17", bty="n")
box(col="gray17")
abline(lm(y2~x), lty="dashed", col="#ef3119", lwd=1.4)
```

- Line of best fit is steeper in the first or second scatterplot?

- Is correlation higher in the first or second scatterplot?

\vspace{-2cm}

## Correlation Coefficient
```{r, cache=FALSE,  fig.width=6, fig.height=6, echo=F, out.width = '40%', fig.align='center'}
par(mfrow = c(1,1),
    oma = c(0,0,0,0) + 0.1, # bottom, left, top, right # outer margins
    mar = c(3,3,1,1) + 0.1, # margin, default: c(5,4,4,2) + 0.1
    mgp = c(1,.5,0), # margin line for axis title, default: c(3,1,0)
    cex=0.8, adj=1, # 1: right-justified
    yaxs="r", xaxs="r") # axis choice: "i" internal, "r" regular
set.seed(1234)
x <- rnorm(1000, mean=0, sd=10)
y <- (x)^2 + rnorm(1000, mean=5, sd=7)
plot(x,y, col="#c7c7c7", pch=19, yaxt="n", xaxt="n", cex.lab=1.5, col.lab="gray17", bty="n", xlab="", ylab="", xlim=c(-35,35))
box(col="gray17")
```

- cor(X,Y) $\approx$ \pause 0

- Does this mean that there is no relationship between the two variables? No. [Check the dino!](https://github.com/umbertomig/POLI30Dpublic/blob/main/images/dino.gif)

\vspace{-1cm}


## Correlation does not necessarily imply causation

\vspace{.6cm}
\begin{center}
\includegraphics[height=.70\textheight]{../../images/correlation_not_causation.jpg}
\end{center}

## Correlation does not necessarily imply causation

- Just because two variables have a strong correlation does not mean that changes in one variable cause changes in the other
\vspace{.2cm}

- Example: *reading* and *math* are highly correlated.
    + Does that mean that if you study *math* you learn *reading*?!

\vspace{-2cm}

## Correlation does not necessarily imply causation

\begin{center} \includegraphics[width=1\textwidth]{../../images/eng.png}\end{center}

## Correlation does not necessarily imply causation

\begin{center} \includegraphics[width=1\textwidth]{../../images/kentucky.png}\end{center}

- More on this later in the semester!

## Summary

\vspace{1cm}

* **Today's Class:**
  + Exploring the Relationship Between Two Variables
    + Scatterplots
    + Correlations

\vspace{1cm}

* Next class:
  + Prediction and Linear Regression
  
# Questions?

# See you in the next class!