---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 08 | R Plots and R Data Analysis IV}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: '../../latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, echo=FALSE, results='hide', include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::knit_hooks$set(myscriptsize = function(before, options, envir) { 
  if (before) { 
    return("\\scriptsize") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(myscriptsize = TRUE)
knitr::opts_chunk$set(collapse = TRUE)
```

## Before we start

**Announcements:**

- GitHub page: [**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: Check with instructors for an alternative link.

## Before we start

**Announcements: Final Exam**

- The best way to study for the **final exam** is to:
  1. Revise the lectures' content from lecture one until the last lecture. All will be on.
  2. Make sure you understand how to run the code and how to interpret results.
  3. Revise the content from the homework. They are a good clue regarding the format of the exam.
  4. If you cannot do it, then explain with words how you would do it. Explain in detail. 
    + This helps us to give you partial credit.

## Before we start

**Recap:** In the Lab sessions so far, you learned:

- How to install R and R Studio on your computer.
- How to do basic and advanced operations with vectors and data frames.
- How to install packages and work with R Markdown.
- How to create plots and how to do data analysis.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 08}}}\normalsize\\\\
- Group-by and summarize\\
- A bit more recoding\\
- Dealing with missing data\\
- Extract random samples from data\\
- Playing with random variables\\
- Full summary of a regression\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# Getting started

## Getting started

- To get started, we need to load the datasets we will need in the lab.

- We also need to load the `tidyverse` package, which has all the R functions we use.

\vspace{0.5cm}

```{r}
library(tidyverse)
```

## Getting started - Education expenditure data

```{r, collapse=TRUE}
educexp <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/educexp.csv")
head(educexp)
```

\vspace{-1cm}

## Getting started - Chile survey data

```{r, collapse=TRUE}
chile <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/survchileComplete.csv")
head(chile)
```

\vspace{-1cm}

## Getting started - Voting

```{r, collapse=TRUE}
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
head(voting)
```

\vspace{-1cm}

# Group-by and Summarize

## Group-by and Summarize

- Suppose we want to find the average age by region.

- This is pretty straightforward when using group-by and summarize:
  + First, we group our results by the `region` 
  + Then, we summarize the `age`

- It will create one average (or whatever stat we ask for) for each region group.

- Syntax:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
dat \%>\% group\_by(groupvar) \%>\% 

\ \ summarize(stat1 = calcs(vars1), etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

\vspace{-1cm}

## Group-by and Summarize

- What operations are available?

\footnotesize

|  **Function**  |             **Operation**             |
|:--------------:|---------------------------------------|
|    `first()`   | First value of a vector               |
|    `last()`    | Last value of a vector                |
|    `nth()`     | Nth value of a vector                 |
|      `n()`     | Number of values in a vector          |
| `n_distinct()` | Number of distinct values in a vector |
|     `min()`    | Minimum value in a vector             |
|     `max()`    | Maximum value in a vector             |
|    `mean()`    | Mean of a vector                      |
|   `median()`   | Median of a vector                    |
|     `var()`    | Variance of a vector                  |
|     `sd()`     | Standard deviation of a vector        |

\normalsize

\vspace{-1cm}

## Group-by and Summarize

- Example: Find the maximum age by region in the Chile Survey data.

\vspace{0.5cm}

```{r, collapse=TRUE}
chile %>% group_by(region) %>%
  summarize(maximumage = max(age))
```

- Note the `NA`. We are going to learn how to deal with those today.

\vspace{-1cm}

## Group-by and Summarize

- Suppose you want to check the vote of the oldest person in each region.

\vspace{0.5cm}

```{r}
chile %>% group_by(region) %>% arrange(desc(age)) %>%
  summarize(voteoldest = first(vote), 
            ageoldest = first(age))
```

- `Y` stands for a vote for Pinochet, and `U` stands for undecided.

- **Question**: How about the vote of the youngest person in each of the regions?

\vspace{-1cm}

# Recoding variables

## Recoding variables

- Create a binary variable that is one when the person has `PS` schooling (some college or more).

\vspace{0.5cm}

```{r, collapse=TRUE}
chile2 <- chile %>%
  mutate(postsec = ifelse(education == 'PS', 1, 0))
chile2 %>% select(education, postsec) %>% head(4)
```

- When using `ifelse`, the syntax is:

> `ifelse(test, val_if_T, val_if_F)`

- Note that we work with vectors in the logical test.

\vspace{-1cm}

## Recoding variables

- Create a binary variable that is one when the person is older than 40.

\vspace{0.5cm}

```{r, collapse=TRUE}
chile3 <- chile %>%
  mutate(olderthan40 = ifelse(age > 40, 1, 0))
chile3 %>% select(age, olderthan40) %>% head(4)
```

- **Your turn:** What is the proportion of people over 40 years old by region?

\vspace{-1cm}

## Recoding variables

- Let us say you want to create a binary indicator of whether the person lives in the `North` or `Central` regions.

- `ifelse()` with `%in%` gets this done:

\vspace{0.5cm}

```{r, collapse=TRUE}
chile4 <- chile %>%
  mutate(NCind = ifelse(region %in% c('N', 'C'), 1, 0))
chile4 %>% count(NCind)
```

- Note the `%in%` operator. This is the operator of choice when we make two or more comparisons!

- **Your turn:** Adapt the code above to add also the South region.

\vspace{-1cm}

## Recoding variables

- Suppose we want to recode a continuous variable into a discrete one.

- Suppose we want to recode age into three groups:
  + From youngest to 30 years old
  + From 31 to 60 years old
  + Older than 60 years old

- To do that, we use the function `cut`. We do three things:
  + `labels = c('lab1', 'lab2, ...)`
  + `breaks = c(-Inf, break1, break2, ..., Inf)`
  + `right = (T or F)`: Add the right break to it?!

\vspace{-1cm}

## Recoding variables

- It is simple to use:

\vspace{0.5cm}

```{r, collapse=TRUE}
chile5 <- chile %>%
  mutate(agecat = 
           cut(age, breaks = c(-Inf, 30, 60, Inf), 
               labels = c('<= 30', '31 to 60', '> 60'), 
               right = T))
chile5 %>% count(agecat)
```

- Note `-Inf` and `Inf`. They stand for $-\infty$ and $\infty$ \smiley

\vspace{-1cm}

# Missing Values

## Missing Values

- Note this `NA` that keeps popping up. This is the way we tell R we have missing data.

- Missing data stands for data that we have no idea is
  + It could be errors in typing up the data
  + It could be that you do not know
  + It could be that a respondent stopped the interview
  + It could be that the person refused to answer a question

- For all these reasons, the key is to understand that it may mess up our analysis.

\vspace{-1cm}

## Missing Data

\vspace{0.5cm}

```{r, collapse=TRUE}
chile %>%
  group_by(region) %>%
  summarize(avgsquo = mean(statusquo), nobs = n())
```

- Note the `NA`s. These are the missing values.

\vspace{-1cm}

## Missing Data

- Most functions in R have a way to deal with it. In `mean`, we add the `na.rm = TRUE` to fix:

\vspace{0.5cm}

```{r, collapse=TRUE}
chile %>%
  group_by(region) %>%
  summarize(avgsquo = mean(statusquo, na.rm = T), nobs = n())
```

\vspace{-1cm}

## Missing Data

- We can remove missing using `na.omit`. It removes the missing values and returns a *clean* dataset.

\vspace{0.5cm}

```{r, collapse=TRUE}
chile %>% select(region, statusquo) %>%
  na.omit() %>% group_by(region) %>%
  summarize(avgsquo = mean(statusquo), nobs = n())
```

- The complete dataset has 2700 observations. 

- After removing the missing in `region` and `statusquo`, it has `r chile %>% select(region, statusquo) %>% na.omit() %>% nrow()` observations.

\vspace{-1cm}

## Missing Data

- When you use `na.omit`, you end up with a smaller dataset.
  + This is the way to go if you do not need the removed cases.

- When you use `na.rm = T`, the dataset remains the same:
  + Good, since missingness can be different in different variables.
  + But sometimes, the function has a different pattern. This works for `mean`, but not for `cor`.
  + It works most of the time, though.
  
# Extracting random samples from data

## Random samples

- Sometimes, we need to extract random samples from a dataset.

- Examples: 
  + Suppose you have the Census in your computer and want to extract random people to survey.
  + Suppose you have a dataset of all students and want to extract a representative sample to run a survey.
  + Suppose you are working with a large dataset, but your computer is old.

- In all these situations, you may extract a sample from your data and work with this sample.

\vspace{-1cm}

## Random samples

- Extract a 10% sample of the data, **without** replacement.

\vspace{0.5cm}

```{r, collapse=TRUE, myscriptsize = TRUE}
set.seed(123456) # change here for a different result
educexp10pct <- 
  educexp %>% sample_frac(0.1, replace = F)
educexp10pct %>% head()
educexp10pct %>% dim()
```

- As you can see, the sample has 10% of the cases or 5 cases.

\vspace{-1cm}

## Random samples

- Extract a 10% sample of the data, **with** replacement.

\vspace{0.5cm}

```{r, collapse=TRUE}
educexp10pctwr <- 
  educexp %>% sample_frac(0.1, replace = T)
educexp10pctwr %>% head()
educexp10pctwr %>% dim()
```

- With replacement, it draws `VT` (Vermont) twice.

\vspace{-1cm}

## Random samples

- Extract a 10-case sample of the data, **without** replacement.

\vspace{0.5cm}

```{r, collapse=TRUE}
educexp10case <- 
  educexp %>% sample_n(10, replace = F)
educexp10case %>% head()
educexp10case %>% dim()
```

- As you can see, the sample has precisely 10 cases.

\vspace{-1cm}

## Random samples

- Extract a 10-case sample of the data, **with** replacement.

\vspace{0.5cm}

```{r, collapse=TRUE}
educexp10casewr <- 
  educexp %>% sample_n(10, replace = T)
educexp10casewr %>% tail(10)
```

- We were lucky that no state had been drawn twice.

\vspace{-1cm}

# Playing with random variables

## Creating Random Bernoulli

- The function `rbinom` gets: Number of cases, $1$ (zeros or ones), and `prob = p`.

\vspace{0.5cm}

```{r, collapse = TRUE, fig.height = 2.2, fig.width = 5}
rv <- rbinom(1000, 1, prob = 0.3)
ggplot(data = data.frame(rv)) + geom_bar(aes(x = rv))
```

\vspace{-1cm}

## Creating Random Bernoulli

\vspace{0.5cm}

- **Your turn:** Create a variable:
\vspace{0.5cm}
  + 10000 observations with probability $p = 0.23$
\vspace{0.5cm}
  + 2000 observations with probability $p = 0.95$
\vspace{0.5cm}
  + 500 observations with probability $p = 0.41$

\vspace{0.5cm}

- Create a histogram of each of the cooked variables.

\vspace{-1cm}

## Creating Random Normal Variables

- The function `rnorm` gets: Number of cases, `mean`, and standard deviation (`sd`).

\vspace{0.5cm}

```{r, collapse = TRUE, fig.height = 2, fig.width = 5}
rv <- rnorm(1000, mean = 33, sd = 7)
ggplot(data = data.frame(rv)) + 
  geom_histogram(aes(x = rv), bins = 20, color = 'white') + 
  theme_minimal()
```

\vspace{-1cm}

## Creating Random Normal Variables

\vspace{0.5cm}

- **Your turn:** Create a variable:
\vspace{0.5cm}
  + 10000 observations with mean $3$ and standard deviation $5$
\vspace{0.5cm}
  + 2000 observations with mean $35$ and standard deviation $10$
\vspace{0.5cm}
  + 500 observations with mean $0$ and standard deviation $1$

\vspace{0.5cm}

- Create a histogram of each of the cooked variables.

\vspace{-1cm}

# Checking the full summary of a regression

## Regression summary
\vspace{0.5cm}
- We can create a summary of a regression. 
\vspace{0.5cm}
- You must wrap the `lm` function around `summary`.
\vspace{0.5cm}
- Check it out for the voting experiment and the expenditure in education in the 1970 US states.
\vspace{0.5cm}
- Couple of things to note:
  1. Lots of statistics. You will learn about the most relevant in class.
  2. You can compute $R^2$ now for models with more than two variables.

\vspace{-1cm}

## Regression summary

```{r, collapse=TRUE, myscriptsize = TRUE}
summary(lm(voted ~ message, data = voting))
```

\vspace{-1cm}

## Regression summary

```{r, collapse=TRUE, myscriptsize = TRUE}
summary(lm(education ~ income + young + urban, data = educexp))
```

\vspace{-1cm}

## Regression summary

- For a prettier table, use the `stargazer` package.
\vspace{0.5cm}
- Stargazer has lots of options and makes plots very pretty. The only one you need to remember: 
\vspace{0.5cm}
  + If your R Markdown is a PDF, use `type = 'latex'`
  \vspace{0.5cm}
  + If your R Markdown is an HTML, use `type = 'html'`
\vspace{0.5cm}
- Also add the `results = 'asis'` to your code chunk, otherwise, it will look weird.

\vspace{-1cm}

## Regression summary

- To install stargazer:
\vspace{0.5cm}
  1. Install `stargazer`
\vspace{0.5cm}
  2. Load `stargazer`

\vspace{0.5cm}

```{r}
library(stargazer)
```

\vspace{-1cm}

## Regression summary

\vspace{0.5cm}

```{r, results='asis'}
mod <- lm(voted ~ message, data = voting)
stargazer(mod, type = 'latex', header = F, float = F,
          font.size = 'scriptsize', no.space = F, 
          dep.var.labels = 'Turnout', style = 'qje',
          covariate.labels = 'Peer-pressure message')
```

\vspace{-1cm}

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Today's Lab}}}\normalsize\\
- Group-by and summarize\\
- A bit more recoding\\
- Dealing with missing data\\
- Extract random samples from data\\
- Playing with random variables\\
- Full summary of a regression\\\\
\multicolumn{1}{c}{\large\concept{\textbf{Next Lab}}}\normalsize\\
- Cool things you can do with R\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}
  
# Questions?

# See you in the next lab!