---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 07 | R Plots and R Data Analysis III}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: '../../latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, echo=FALSE, results='hide', include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(myfntsize = TRUE)
library(tidyverse)
```

## Before we start

**Announcements:**

- GitHub page: [**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: The link in the slides needs to be fixed. Check with instructors for an alternative link.

## Before we start

**Recap:** In the Lab sessions, you learned:

- How to install R and R Studio on your computer.
- How to do basic and advanced operations with vectors and data frames.
- How to install packages and work with R Markdown.
- How to create plots and data viz.
- How to do data analysis.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 07}}}\normalsize\\\\
- Selecting variables\\
- Filtering cases in the dataset\\
- Grouping and summarizing data\\
- Visualizing correlations\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# Getting started

## Getting started

- To get started, we need to load the datasets we will need in the lab.

- We also need to load the `tidyverse` package, which has all the R functions we use.

```{r}
library(tidyverse)
```

## Getting started - Education expenditure data

```{r}
educexp <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/educexp.csv")
head(educexp)
```

## Getting started - Chile survey data

```{r}
chilesurv <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/survchile.csv")
head(chilesurv)
```

## Getting started - Voting

```{r}
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
head(voting)
```

# The `tidyverse` package

## The `tidyverse` package

- The `tidyverse` package is an eco-system for data analytics.

- It has packages for processing, plot, and wrangle data.

- We have seen the package for plotting: [`ggplot2`](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf).

- But there are other important packages in the `tidyverse` constellation.

- And in the links we provide URLs for cheat sheets.

\vspace{-1cm}

## The `tidyverse` package

Other imporatant packages:

- [`forcats`](https://posit.co/wp-content/uploads/2022/10/factors-1.pdf): Package to deal with categorical variables.

- [`stringr`](https://posit.co/wp-content/uploads/2022/10/strings-1.pdf): Package to deal with strings and texts in R.

- [`purrr`](https://posit.co/wp-content/uploads/2022/10/purrr.pdf): Package to work with functions and lists.

- [`tidyr`](https://posit.co/wp-content/uploads/2022/10/tidyr.pdf): Package to clean up and reshape datasets.

- [`dplyr`](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf): Package to manipulate (wrangling) datasets.

We are going to do some `dplyr` today!

\vspace{-1cm}

## `dplyr`

- `dplyr` is the package on `tidyverse` for data wrangling.

- Data wrangling is a fancy way to say that you manipulate your dataset until it is in shape for analysis.

- Most of the time, we collect data from the internet, documents, or even old archives.

- We then need to do some *wrangling* to ensure it is ready to extract the statistics of interest.

\vspace{-1cm}

## `dplyr`

`dplyr` is based on a bunch of verbs that do what we intuitively think each of them should do:

- Variable/column:
    + `select`: Select variables
    + `rename`: Rename variables
    + `mutate`: Mutate variables
- Observation/row:
    + `filter`: Filter dataset
    + `arrange`: Arrange observations
- Get summaries:
    + `summarize`: Summarize the dataset
    + `group_by`: Group summary by a variable

\vspace{-1cm}

# Selecting Variables (column operations)

## Selecting Variables (column)

Suppose we want to select only the numeric variables in the `educexp` dataset. We do the following:

```{r, collapse=TRUE}
educ2 <- select(educexp, education, income, young, urban)
head(educ2, 2)
```

- The syntax is: 

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
newdat <- select(olddat, v1, ..., vn)
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

\vspace{-1cm}

## Selecting Variables (column)

Or a more compact notation for this uses *pipes*:

```{r, collapse=TRUE}
educ3 <- educexp %>%
  select(education, income, young, urban)
head(educ3, 2)
```

- Note that the *pipe* operator (`%>%`) works like a composite function: takes what is in the left and passes through to the right.

- The first thing that the select is looking for is the dataset. The *pipe* passes the dataset, so you do not need to mention it.

\vspace{-1cm}

## Selecting Variables (column)

And the methods we can apply in the `select` are the following:

| Method             | Effect                                          |
|--------------------|-------------------------------------------------|
| v1, v2, v3 (etc)   | Select given variables                          |
| starts_with('xyz') | Select starting with `xyz`                      |
| ends_with('xyz')   | Select ending with `xyz`                        |
| contains('xyz')    | Select variables that have `xyz` in their names |
| vk:vn              | All variables between `vk` and `vn`             |
| -(vk:vn)           | All but the variables between `vk` and `vn`                            |

We can even rename variables using `select`.

## Selecting Variables (column)

Example 1 (same selection as before):

```{r, collapse=TRUE}
educ4 <- educexp %>% select(-states)
head(educ4, 3)
```

Example 2 (in the Chile dataset now):

```{r, collapse=TRUE}
chile2 <- chilesurv %>% select(starts_with('vote'))
head(chile2, 3)
```

## Rename

We can change the names of the variables using `rename`. The basic syntax is:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
newdat <- olddat %>%
  rename(newname1 = oldname1,
         newname2 = oldname2,
         etc)
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

Example:

```{r, myfntsize = T, collapse=TRUE}
chile3 <- chilesurv %>% rename(votebinary = voteYES)
head(chile3, 2)
```

\vspace{-1cm}

## Pipeing like a boss

You can *pipe* as many commands as you want. Example:

```{r, myfntsize = T, collapse=TRUE}
chile4 <- chilesurv %>%
  select(starts_with('vote')) %>%
  rename(votebinary = voteYES)
head(chile4, 2)
```

- Note the *chain* pipe!

\vspace{-1cm}

# Filtering Rows

## Filtering Rows

- Filtering the data means subset the dataset based on a logical condition.
  + Example: In the voting pressure experiment, we may want to study the effect of messaging young people.
  + Reason: Younger people may not be as sensitive to peer pressure.

```{r, myfntsize = TRUE, collapse=TRUE}
voting2 <- voting %>% filter(birth >= 1975)
head(voting2, 4)
```

\vspace{-1cm}

## Filtering Rows

- Syntax:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
newdat <- filter(olddat, condition1, condition2, etc)

or

newdat <- olddat \%>\% filter(condition1, condition2, etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- Note that you may or may not use the pipe operator. But we can all agree that it makes the code much cleaner.

## Filtering rows

* The filter operators are the following:

| **Operator** | **Meaning**                                |
|----------|----------------------------------------|
| `<` or `<=` | Smaller than or smaller than or equal |
| `>` or `>=` | Greater than or greater than or equal |
| `==`       | Equal                                  |
| `!=`       | Different                              |
| `!`        | Negation (turns a TRUE into a FALSE)                              |
| `|`        | Or                                     |
| `&`        | And                                    |

\vspace{-1cm}

## Arrange

- We can sort the dataset by variable content according to the needs of our analysis.

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
newdat <- olddat \%>\% arrange(var1, desc(var2), etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- The `desc()` command sorts descending (highest to the lowest). The default is to sort ascending.

\vspace{-1cm}

## Arrange

- Example: Sorting by education expenditure (high to low):

```{r, myfntsize = T, collapse=TRUE}
educexp %>% arrange(desc(education)) %>% head()
```
\vspace{-1cm}

# Mutating (altering the content)

## Mutating

- We frequently need to operate with our variables. For example, suppose that you want to apply a log function to `income`.

- `log` has the advantage of transforming large changes into smaller ones.

- It also has the advantage of interpreting regression models. Check [out this UofV Library entry](https://data.library.virginia.edu/interpreting-log-transformations-in-a-linear-model/).

- Syntax:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
newdat <- olddat \%>\% mutate(vnew = calcs(vold), etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

\vspace{-1cm}

## Mutating

Example: Log of `education` and the sum of variables:

```{r}
educ5 <- educexp %>% 
  mutate(logeduc = log(education), 
         sum_iyu = income + young + urban)
head(educ5, 4)
```

**Your turn:** Adapt this code to arrange the dataset in terms of `logeduc`.

\vspace{-1cm}

# Grouping(-by) and Summarizing

## Summarize

- The `summarize` function is helpful when computing summary statistics in the whole sample.

- Syntax:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
olddat \%>\% summarize(stat1 = calcs(vars1), stat2 = calcs(vars2), etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- Note that we only save the results if we need them! The idea is that those summary statistics you compute to check, not save.

\vspace{-1cm}

## Summarize

- Example: Suppose we want to compute the mean, standard deviation, and the number of observations for `education` and `income`.

```{r, myfntsize = T, collapse=TRUE}
educexp %>%
  summarize(avgeduc = mean(education), 
            stdeveduc = sd(education), neduc = n(),
            avginc = mean(income), 
            stdevinc = sd(income), ninc = n())
```

- Besides the number of observations, these are pretty informative statistics.

\vspace{-1cm}

## Group-by and Summarize

- But we may want the summary by a group. For example, to compute the *differences-in-means* estimator, we need:
  + The average in the treatment group
  + The average in the control group.
  
- Group-by can help us here: 
  + First, we group our results by the treatment status. 
  + Then, we summarize. 

- It will create one summary for each group.

- Syntax:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
dat \%>\% group\_by(groupvar) \%>\% 

\ \ summarize(stat1 = calcs(vars1), etc)
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

\vspace{-1cm}

## Group-by and Summarize

- Example:

```{r, collapse=TRUE}
voting %>% group_by(message) %>%
  summarize(avgvote = mean(voted) * 100, ncases = n())
```

- And the differences-in-means estimator is then straightforward to compute.

- Note that we also saved the number of cases in each treatment status. 
  + It makes more sense: Number of control and treatment groups cases.

\vspace{-1cm}

## Group-by and Summarize

**Your turn:** Adapt this example here to compute the percentage of voters in the treatment and the control group, filtering by younger voters (born on or after 1975).

- Example:

```
voting %>% 
  group_by(message) %>%
  summarize(avgvote = mean(voted) * 100, 
            ncases = n())
```

**Question:** Does the treatment effect increase or decrease? Why do you think that is the case?

\vspace{-1cm}

## Detour: `tibble` versus `data.frame`

- When using `dplyr` functions such as the ones we have used in here, `tidyverse` sometimes saves the results in `tibble`s.

- `tibble`s are pretty much the same as `data.frame`s.

- To go back to `data.frame`, you should do the following: `dfdat <- data.frame(tibbledat)`.

- Or you can add a `%>% data.frame()` to the end of your pipeing. For this class, it does not matter which one you use.

- If you want to check about `tibble`s, [here](https://tibble.tidyverse.org) is a good source.

\vspace{-1cm}

# Plotting Correlations

## `ggcorrplot` and `GGally`

- These packages are helpful to add some more spice to our plots.

- For example, `ggcorrplot` is good for plotting correlations of multiple variables.
  + The command `cor` computes two-by-two correlations for all the combinations of variables when more the two vars.

- Installing packages: Run this code in your console:

```
install.packages('ggcorrplot')
install.packages('GGally')
```

- \alert{Warning:} You should ***never*** install packages on R Markdown!

\vspace{-1cm}

## `ggcorrplot` and `GGally`

- Now, to load both packages:

```{r}
library('ggcorrplot')
library('GGally')
```

- And we now have all the functionalities of these packages at our disposal!

\vspace{-1cm}

***
```{r, fig.height=3, fig.width=4.5, fig.align='center', myfntsize = TRUE, collapse=TRUE}
ggpairs(educexp %>% select(-states), aes(alpha = 0.4)) + 
  theme(text = element_text(size = 10), 
        axis.text.x = element_text(angle = 90) ) 
```
\vspace{-0.5cm}

## Correlations

- And the correlation between these variables is:

```{r, collapse=TRUE}
educexp %>% select(-states) %>% cor()
```

- The correlation between the variable and itself is always one.

- The correlation is a **`symmetric matrix`**: 
  + The correlation between `income` and `education` is the same as the correlation between `education` and `income`.

## Correlations

Using squares:

```{r, fig.height=2.5, fig.width=4, fig.align='center', myfntsize = TRUE, collapse=TRUE}
ggcorrplot(educexp %>% select(-states) %>% cor()) 
```

\vspace{-2cm}

## Correlations

And using circles:

```{r, fig.height=2.4, fig.width=4, fig.align='center', myfntsize = TRUE, collapse=TRUE}
ggcorrplot(educexp %>% select(-states) %>% cor(), 
           method = 'circle') 
```

\vspace{-2cm}

## `ggcorrplot` and `GGally`

- Note that both commands use **only numeric variables.**

- Another thing to note is that you can do correlation plots using `GGally`.
  + Function name: `ggcorr`. The syntax is different, but it also produces a nice plot.
  
- Syntax for `ggpairs`:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{semiverbatim}
dat \%>\% select(numericvars) \%>\% ggpairs()
\end{semiverbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

- Syntax for `ggcorrplot`:

\vspace{0.2cm}
\footnotesize
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{verbatim}
dat %>% select(numericvars) %>% 
  cor() %>% ggcorrplot()
\end{verbatim}
\end{mdframed}
\vspace{.2cm}
\normalsize

\vspace{-1cm}

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Today's Lab}}}\normalsize\\
- Selecting variables\\
- Filtering cases in the dataset\\
- Grouping and summarizing data\\
- Visualizing correlations\\\\
\multicolumn{1}{c}{\large\concept{\textbf{Next Lab}}}\normalsize\\
- More data wrangling\\
- One more plot\\
- Dealing with missing data\\
- Lots of analysis\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}
  
# Questions?

# See you in the next lab!