---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 09 | Prediction II}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, include=FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- GitHub page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: Not sure what the link is. Ask your TA!

- Note to self: Turn on the mic!

## Before we start

**Recap:** We learned:

- The definitions of theory, scientific theory, and hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.
- Sampling, descriptive statistics, and descriptive plots for one variable.
- Correlation between two continuous variables.
- Prediction of a non-binary variable.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.85\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\\\
- Prediction and Linear Regression\\
- Example with Binary Outcome Variable: \\
\,\, Using status quo Scores to Predict Probability\\ 
\,\,\, of Supporting a Dictator\\[.3cm]
\,\,\,\, 1. Load and explore data\\
\,\,\,\, 2. Identify X and Y\\
\,\,\,\, 3. What is the relationship between X and Y?\\
\,\,\,\,\,\,\,\,\,\,\,\,         - Create scatter plot \\
\,\,\,\,\,\,\,\,\,\,\,\,         - Calculate correlation\\
\,\,\,\, 4. Fit a linear model using the least squares method\\
\,\,\,\, 5. Interpret coefficients\\
\,\,\,\, 6. Make predictions\\
\,\,\,\, 7. Measure how well the model fits the data
\end{tabular}
\end{center}
\vspace{.2cm}
\end{mdframed}
\end{minipage}
\end{center}

## Predicting Support for a Dictator

\footnotesize

- In 1988, FLACSO ran a survey to estimate the support for [Augusto Pinochet](https://en.wikipedia.org/wiki/Augusto_Pinochet) in Chile.

\vspace{-0.1cm}

- This survey was conducted on the eve of a referendum that could have ousted Pinochet.

\vspace{-0.1cm}

| **variable** | **meaning** |
|---|---|
| `statusquo` | Scale with status-quo evaluation. Roughly from -5 to 5. |
| `vote` | Declared vote in the upcoming referendum. |
| `voteYES` | 1 means vote for Pinochet, and 0 means vote against it. |

\vspace{-0.5cm}

- We will study whether a person satisfied with the status quo would tend to favor Pinochet in the plebiscite.

\normalsize

\vspace{-1cm}

## Step 1: Load and explore data

```{r, collapse=TRUE, comment='##'}
survchile <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/survchile.csv")
head(survchile)
```
\vspace{-.1cm}

- What is the unit of observation?

- For each variable: type and unit of measurement?

- Substantively interpret the first observation.

\vspace{-1cm}

## Step 2: Identify the Dependent and Independent Variables

- The \concept{predictor (X)}\ is the variable we want to use to predict the outcome (Y).

- The \concept{target (Y)}\ is the variable that we want to predict.

- What are they?

```{r, fig.width=6, fig.height=2, echo=FALSE}
p1 <- ggplot(data = survchile, aes(x = statusquo)) + 
  geom_histogram(bins = 7, color = 'black', fill = 'lightblue', alpha = 0.4) + labs(title = 'Status Quo Scale', y = '') + 
  theme_minimal()
p2 <- ggplot(data = survchile, aes(x = voteYES)) + 
  geom_histogram(bins = 7, color = 'black', fill = 'lightblue', alpha = 0.4) + labs(title = 'Intention to vote YES\nin the plebiscite', y = '') + 
  theme_minimal()
grid.arrange(p1, p2, nrow = 1)
```

## Step 2: Identify the Dependent and Independent Variables

- What type of variable is *voteYES*?
  + Binary

- How would you compute the proportion of intended `Yes` votes?
  + By computing the mean of *voteYES*
  
  + Since *voteYES* is a binary variable, its mean should be interpreted as the proportion of the observations that have the characteristic identified by the variable

## Step 2: Identify the Dependent and Independent Variables

- Code to compute the mean of *voteYES*
  + Answer:
```{r, collapse=TRUE, comment='##'}
mean(survchile$voteYES)
```
- Interpretation?
  + Close to `r round(mean(survchile$voteYES), 4)*100`% of people responded that they intended to support Pinochet in the upcoming plebiscite.
  + RECALL: You need to multiply the output by 100

## Step 2: Identify the Dependent and Independent Variables

- Since Y is binary:
\vspace{.1cm}
  + unit of measurement of $\overline{Y}$?
    + \% (after x 100)
  + unit of measurement of $\widehat{\beta}_0$?
    + \% (after x 100)
  + unit of measurement of $\widehat{Y}$?
    + \% (after x 100)
  + unit of measurement of $\triangle\overline{Y}$?
    + p.p. (after x 100)
  + unit of measurement of $\widehat{\beta}_1$?
    + p.p. (after x 100)
  + unit of measurement of $\triangle\widehat{Y}$?
    + p.p. (after x 100)
    
## Step 3: What is the relationship between X and Y?

- Create \concept{scatter plot}\ to visualize the relationship between *statusquo* and *voteYES*.

```{r, fig.width=6, fig.height=2.5, echo=FALSE}
ggplot(data = survchile, aes(x = statusquo, y = voteYES)) + 
  geom_point(fill = 'lightblue', alpha = 0.4) + 
  labs(title = '', y = 'Intention to vote YES in the plebiscite', x = 'Status-Quo Scale') + 
  theme_minimal()
```

- It is hard to see the y-axis variation. We add a little `jitter` on y, then.

\vspace{-1cm}

## Step 3: What is the relationship between X and Y?

```{r, fig.width=6, fig.height=3, echo=FALSE}
ggplot(data = survchile, aes(x = statusquo, y = voteYES)) + 
  geom_jitter(fill = 'lightblue', alpha = 0.5, height = 0.4, width = 0) + 
  labs(title = '', y = 'Intention to vote YES\nin the plebiscite', x = 'Status-Quo Scale') + 
  theme_minimal()
```

\vspace{-0.5cm}

- What does each dot represent?

- Does the relationship look positive or negative?

- Does the relationship look weekly or strongly linear?

\vspace{-1cm}

## Step 3: What is the relationship between X and Y?

- Calculate \concept{correlation}\ to measure direction and strength of linear association between *statusquo* and *voteYES* \pause

```{r, collapse=TRUE, comment='##'}
cor(survchile$statusquo, survchile$voteYES)
```
- We find a strong positive correlation

- Are we surprised by this?

## Step 4: Fit a linear model using the least squares method

- R function to fit a linear model: \fu{lm()}

```{r, echo=FALSE}
mod <- lm(voteYES ~ statusquo, data = survchile)
```

```{r, collapse=TRUE, comment='##'}
lm(voteYES ~ statusquo, data = survchile)
```

- $\widehat{\beta}_0$ = `r round(mod$coef[1], digits=2)` and $\widehat{\beta}_1$ = `r round(mod$coef[2], digits=2)`

- The fitted line is $\widehat{Y}$ = `r round(mod$coef[1], digits=2)` + `r round(mod$coef[2], digits=2)` $X$

- More specifically, it is $\widehat{\textrm{voteYES}}$ = `r round(mod$coef[1], digits=2)` + `r round(mod$coef[2], digits=2)` \textit{statusquo}

\vspace{-1cm}

## Step 4: Fit a linear model using the least squares method

- R function to add a fitted line to scatter plot: \fu{geom\_smooth()} 

```{r, fig.width=6, fig.height=2.2, mytiny = TRUE}
ggplot(data = survchile, aes(x = statusquo, y = voteYES)) + 
  geom_jitter(fill = 'lightblue', alpha = 0.5, height = 0.4, width = 0) + 
  labs(title = '', y = 'Intention to vote YES in the plebiscite', x = 'Status-Quo Scale') + 
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1) +
  theme_minimal()
```

\vspace{-1cm}

## Step 5: Interpretation of Coefficients

\small

- **Substantive interpretation of $\widehat{\beta}_0$**?
  + Start with the mathematical definition: 
    + $\widehat{\beta}_0$ is the $\widehat{Y}$ when *X*=0
  + Substitute X, Y, and $\widehat{\beta}_0$: 
    + $\widehat{\beta}_0$ = `r round(mod$coef[1], digits=2)` is the $\widehat{\textrm{\textit{voteYES}}}$ when *statusquo*=0
  + Put it in words (using units of measurement): 
    + When a person is neither happy nor sad with things as they are, we predict that her probability of voting YES in the plebiscite is `r round(mod$coef[1], digits=2)*100`\%, on average

- **Unit of measurement of $\widehat{\beta}_0$**? 
    + Same as $\overline{Y}$
    + Since *Y* is binary, $\overline{Y}$ is measured in \%, and so is $\widehat{\beta}_0$ (after x 100)

\normalsize

\vspace{-1cm}

## Step 5: Interpretation of Coefficients

\small

- **Substantive interpretation of $\widehat{\beta}_1$**?
  + Start with the mathematical definition: 
    + $\widehat{\beta}_1$ is the $\triangle \widehat{Y}$ associated with $\triangle X$=1
  + Substitute X, Y, and $\widehat{\beta}_1$: 
    + $\widehat{\beta}_1$ =`r round(mod$coef[2], digits=2)` is the $\triangle \widehat{\textrm{\textit{voteYES}}}$ associated with $\triangle$*statusquo*=1
  + Put it in words (using units of measurement): 
    + Increasing satisfaction with the status quo by 1 point is associated with a predicted increase in the chance of voting YES of `r round(mod$coef[2], digits=2)*100` p.p., on average

- **Unit of measurement of $\widehat{\beta}_1$**?
    + Same as $\triangle \overline{Y}$
    + Since *Y* is binary, $\triangle \overline{Y}$ is measured in p.p., and so is $\widehat{\beta}_1$ (after x 100)

\normalsize

\vspace{-1cm}

## Step 5: Interpretation of Coefficient

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm, innerleftmargin=.5cm,innerrightmargin=.5cm]
\vspace{.3cm} 
\begin{center}\color{concept}\tip{The fitted line is}\color{text}\end{center}
$$\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$$

\begin{tabular}{l}
- $\widehat{\beta}_0$ (beta-zero-hat) is the estimated intercept coefficient\\
\,\,\,\,\,\,\,\,\,\,\,the $\widehat{Y}$ when $X{=}\textrm{0}$\\
\,\,\,\,\,\,\,\,\,\,\,(in same unit of measurement as $\overline Y$)\\\\

- $\widehat{\beta}_1$ (beta-one-hat) is the estimated slope coefficient\\
\,\,\,\,\,\,\,\,\,\,\,the $\triangle \widehat{Y}$ associated with $\triangle X{=}\textrm{1}$\\
\,\,\,\,\,\,\,\,\,\,\,(in the same unit of measurement as $\triangle\overline Y$)
\end{tabular}
\vspace{.2cm}
\end{mdframed}

\vspace{-1cm}

## Step 6: Make predictions

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm, innerleftmargin=.5cm,innerrightmargin=.5cm]
\vspace{.3cm} 
\begin{center}\color{concept}\tip{Using the Fitted Line}

\vspace{-.3cm}
\tip{to Make Predictions}\color{text}\end{center}
\begin{tabular}{ll}
\\
- To predict $\widehat{Y}$ based on X: & \,\,\,\,\,\,\,\,$\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$\\\\
- To predict $\triangle\widehat{Y}$ based on $\triangle$X:&\,\,\,\,\,\,\,\,$\triangle\widehat{Y} = \widehat{\beta}_1 \triangle X$\\\\
\end{tabular}
\vspace{.2cm}
\end{mdframed}
  
## Step 6: Make predictions

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{-.1cm}
\noindent To predict $\widehat{Y}$ based on $X$: \color{titleline} $\widehat{Y} = \widehat{\beta}_0 + \widehat{\beta}_1 X$
\end{mdframed}

- \underline{Example 1}: Imagine a person is unsatisfied with things and evaluates the status quo as -2. What would we predict her chance of favoring Pinochet in the plebiscite?

\vspace{-.9cm}
\begin{eqnarray*}
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \,\,\textrm{\,\,statusquo}\\
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \times \textrm{-2.0} \color{gray} \textrm{\,\, (if statusquo = -2.0)}\\
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*(-2)`} \\
\end{eqnarray*}
\vspace{-1.5cm}

- Answer: If her status quo evaluation is -2.0 points, we would predict that her probability of supporting Pinochet is of `r (round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*(-2))*(100)`\%, on average
- Note: since Y is binary, $\widehat{Y}$ is measured in \% (after x 100)

\vspace{-2cm} 

## Step 6: Make predictions

- \underline{Example 2}: Imagine a person is happy with things and evaluates the status quo as 2. What would we predict her chance of favoring Pinochet in the plebiscite?

\vspace{-.9cm}
\begin{eqnarray*}
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \,\,\textrm{\,\,statusquo}\\
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2)`} + \textrm{`r round(mod$coef[2], digits=2)`} \times \textrm{2.0} \color{gray} \textrm{\,\, (if statusquo = 2.0)}\\
\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*2.0`} \\ 
\end{eqnarray*}
\vspace{-1.5cm}

- Answer: If the person scores 2.0 points on the status quo scale, we would predict that she would vote for Pinochet `r (round(mod$coef[1], digits=2) + round(mod$coef[2], digits=2)*2)*100`\% of the time, on average

\vspace{-1cm}

## Step 6: Make predictions

\begin{mdframed}[linecolor=titleline, linewidth=0.05cm]
\noindent To predict $\triangle \widehat{Y}$ associated with $\triangle X$: \color{titleline} $\triangle \widehat{Y} = \widehat{\beta}_1 \, \triangle \text{X}$
\end{mdframed}

- \underline{Example 3}: If we raise a person's status quo evaluation by three points, how much would we predict that her support for Pinochet would change?

\vspace{-.9cm}
\begin{eqnarray*}
\triangle\widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[2], digits=2)`} \,\,\triangle\textrm{statusquo}\\
\triangle \widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[2], digits=2)`} \times \textrm{3.0} \color{gray} \textrm{\,\, (if $\triangle$statusquo = 3.0)} \color{black} \\ 
\triangle \widehat{\textrm{voteYES}} &=& \textrm{`r round(mod$coef[2], digits=2)*3`} \\
\end{eqnarray*}
\vspace{-1.6cm}

- Answer: An increase of status quo scores of 3 points is associated with a predicted increase in the probability of voting yes in the plebiscite of `r round(mod$coef[2], digits=2)*3*100` p.p., on average.

- Note: Since Y is binary, $\triangle\widehat{Y}$ is in p.p. (after x 100)

\vspace{-2cm}

## Step 7: Measure how well the model fits the data

- How good is the model at making predictions? How well does the model fit the data?

- One way of answering is by calculating \concept{$R^2$}

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=white,linecolor=concept, linewidth=0.05cm]
\concept{$R^2$}\ measures the proportion of the variation in the outcome variable explained by the model
    \begin{itemize}
    \item It ranges from 0 to 1
    \item The higher the $R^2$, the better the model fits the data
    \item In the simple linear model: $R^2 = \textrm{\textit{cor}}(X,Y)^2$
    \item The higher the correlation between X and Y (in absolute terms), the better the model fits the data
    \end{itemize}
\end{mdframed}

\vspace{-1cm}

## Step 7: Measure how well the model fits the data

\small

- When cor(X,Y) = 1 or cor(X,Y) = -1, the relationship between X and Y is perfectly linear.

\vspace{-0.1cm}

- $R^2$ = cor(X,Y)$^2$ = 1, the model explains 100\% of the variation of Y.

\vspace{-0.1cm}

- All prediction errors (vertical distance between the dots and the line) = 0.

\vspace{-0.1cm}

- When cor(X,Y) = 0, the relationship between X and Y is non-linear.

\vspace{-0.1cm}

- $R^2$=cor(X,Y)$^2$ = 0, the model explains 0\% of the variation of Y.

\vspace{-0.1cm}

- The prediction errors (vertical distance between the dots and the line) are large.

\vspace{-1cm}

\normalsize

## Step 7: Measure how well the model fits the data

```{r, fig.width=6, fig.height=2, echo = FALSE}
ggplot(data = survchile, aes(x = statusquo, y = voteYES)) + 
  geom_jitter(fill = 'lightblue', alpha = 0.5, height = 0.4, width = 0) + 
  labs(title = '', y = 'Intention to vote YES\nin the plebiscite', x = 'Status-Quo Scale') + 
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1) +
  theme_minimal()
```

- Let us compute $R^2$
```{r, collapse=TRUE, comment='##'}
cor(survchile$statusquo, survchile$voteYES)^2
```
    
\vspace{-2cm}

## Step 7: Measure how well the model fits the data

- Let us compute $R^2$:

```{r, collapse=TRUE, comment='##'}
cor(survchile$statusquo, survchile$voteYES)^2
```

- Interpretation?

  + It means that the linear model explains `r round(cor(survchile$statusquo, survchile$voteYES)^2, 2)*100`\% of the variation of the outcome variable (*voteYES*)
  
  + **Note:** It does NOT mean that the model is right `r round(cor(survchile$statusquo, survchile$voteYES)^2, 2)*100`\% of the time. 

\vspace{-2cm}

## Step 7: Measure how well the model fits the data

Let us return to the predictive model from the last lecture:

```{r, fig.width=6, fig.height=3, echo = FALSE}
educexp <- read.csv('https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/educexp.csv')
ggplot(data = educexp, aes(x = income, y = education)) + geom_text(aes(label=states), size=2) + 
  labs(title = '', y = 'Education Expenditure', x = 'Income') +
  geom_smooth(formula = 'y ~ x', method = 'lm', se = F, color = 'blue', lwd = 1) + theme_minimal()
```

\vspace{-1cm}

## Step 7: Measure how well the model fits the data

```{r, collapse=TRUE, comment='##'}
cor(educexp$income, educexp$education)^2
```

- Interpretation?
  + It means that the linear model explains `r round(cor(educexp$income, educexp$education)^2, 2)*100`\% of the variation of the outcome variable (*education*)
  + It does NOT mean that the model is right `r round(cor(educexp$income, educexp$education)^2, 2)*100`\% of the time

- **Warnings:**
  1. Only compare $R^2$ between models with the same outcome variable ($Y$)
  2. Some variables are intrinsically harder to predict than others

\vspace{-1cm}

***
\vspace{.8cm}
\begin{mdframed}[linecolor=titleline, linewidth=0.05cm] 
\vspace{.3cm} \color{black}
\tip{Predicting Outcomes Using Linear Models}: 
We look for $X$ variables that are highly correlated with $Y$ because the higher the correlation between $X$ and $Y$ (in absolute terms), the higher the $R^2$ and the better the fitted linear model will usually be at predicting $Y$ using $X$.
\vspace{.2cm}
\end{mdframed}


## Summary

- **Today's Class:**
  + Practice summarizing the relationship between X and Y with a line: \fu{lm()}.
  + Practice interpreting the two estimated coefficients ($\widehat{\beta}_0$ and $\widehat{\beta}_1$) when outcome variable is binary.
  + Practice making predictions with the fitted line: predict $\widehat{Y}$ based on $X$ and predict $\triangle\widehat{Y}$ based on $\triangle X$.
  + Learned how to measure how well the model fits the data with $\textrm{R}^2$.

\vspace{0.2cm}

- **Next class**:
  + Causality with Observational Data
  
\vspace{-1cm}


# Questions?

# See you in the next class!