---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{Professor Umberto Mignozzetti\\(Based on DSS Materials)}\vspace{.3cm}
date: |
    | \large\textbf{Lecture 05 | Does Social Pressure Increase Probability of Turning Out to Vote?}
fontsize: 12pt
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: 'https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r echo=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
```

## Before we start

**Announcements:**

- Quizzes and Participation: On Canvas.

- Github page: [https://github.com/umbertomig/POLI30Dpublic](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: [https://piazza.com/ucsd/winter2023/17221](https://piazza.com/ucsd/winter2023/17221)

## Before we start

**Recap:**

- We learned the definitions of Theory, Scientific Theory, and Hypotheses.
- Data, datasets, variables, and how to compute means.
- Causal effect, treatments, outcomes, and randomization.

**Great job!**

- Do you have any questions about these contents?


***
\begin{center}
\begin{minipage}{0.8\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Today}}}\normalsize\\\\
- Percentage Points\\
- Review: Unit of Measurement of Means\\
- Unit of Measurement of Diffs-in-Means\\\\
- In-Class Exercise: Does Social Pressure \\
\,\,\,Increase the Probability of Turning Out To Vote? \\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

## What is a percentage point? 

Unit of measurement for the arithmetic difference between two percentages:
\begin{center}\concept{\% - \% = p.p.}\end{center}

+ Example: if a candidate's vote share increases from 50\% to 60\%, the vote share increases by
\vspace{-.2cm}
$$\triangle\textrm{vshare} = \textrm{vshare}_{\textrm{final}} - \textrm{vshare}_{\textrm{initial}}  =  \textrm{60\%} -  \textrm{50\%} =  \textrm{10 p.p.}$$
\vspace{-.8cm}

-  Why not 10\%?
    +  What is 10\% of 50\%?  $\textrm{0.10}{\times}\textrm{50}=\textrm{5 p.p.}$
\vspace{-.2cm}
$$\textrm{vshare}_{\textrm{final}} = \textrm{vshare}_{\textrm{initial}} + \triangle \textrm{vshare}  = \textrm{50\%} +  \textrm{5 p.p.} =  \textrm{55\%}$$
\vspace{-1.6cm}

## Unit of Measurement of the Diffs-in-Means Estimator
\begin{mdframed}[backgroundcolor=white,linecolor=titleline, linewidth=0.05cm]
\begin{center} \color{text}
\begin{tikzpicture}
\node [above] at (0,10.3) {{unit of measurement of the difference-in-means estimator}};
\draw (0,10.3)--(0,10.1);
\draw (-2.4,10.1)--(2.4,10.1);
\draw (-2.4,10.1)--(-2.4,10);
\draw (2.4,10.1)--(2.4,10);
\node [below] at (-2.4,10) {{\color{title}if outcome variable}};
\node [below] at (-2.4,9.6) {{\color{title}is non-binary:}};
\node [below] at (-2.4,9.2) {{in the same}};
\node [below] at (-2.4,8.8) {{unit of measurement}};
\node [below] at (-2.4,8.4) {{as the outcome variable}};

\node [below] at (2.4,10) {{\color{title}if outcome variable}};
\node [below] at (2.4,9.6) {{\color{title}is binary:}};
\node [below] at (2.4,9.2) {{in percentage points}};
\node [below] at (2.4,8.8) {{(after multipying}};
\node [below] at (2.4,8.4) {{the result by 100)}};
\end{tikzpicture}
\end{center}
\end{mdframed}

## Does Social Pressure Affect Turnout?
\vspace{.4cm}\begin{center}\includegraphics[height=.6\textheight]{../../images/vote.jpg}\end{center}\vspace{-.3cm}

\footnotesize
(Based on \href{http://www.donaldgreen.com/wp-content/uploads/2015/09/Gerber_Green_Larimer-APSR-2008.pdf}{Alan S. Gerber, Donald P. Green, and Christopher W. Larimer. 2008. ``Social Pressure and Voter Turnout: Evidence from a Large-Scale Field Experiment." \textit{American Political Science Review}, 102 (1): 33-48.})

## Does Social Pressure Affect Turnout?

- We will answer by analyzing data from an experiment where registered voters were randomly assigned to either 
    + (a) receive a message designed to induce social pressure to vote, or
    + (b) receive nothing

\vspace{.2cm}

- The message told that after the election, their neighbors would be informed about their voting decision:
    + What do you think the effect might be?
\vspace{-1cm}

## Does Social Pressure Affect Turnout?

- What do we need to calculate the ATE of receiving the message on the probability of turning out to vote?
\vspace{.1cm}
    + the difference-in-means estimator

\vspace{.2cm}    

- Why does the difference-in-means estimator provide a *valid* estimate of the ATE?
\vspace{.1cm}
    + the data come from a randomized experiment (where treatment was randomly assigned)
    + as a result, treatment and control groups are comparable
    
## Does Social Pressure Affect Turnout?

- In this case, the difference-in-means estimator is:
\vspace{.2cm}
\begin{center}
$\overline{\textrm{\textit{voted}}}_\text{treatment group} - \overline{\textrm{\textit{voted}}}_\text{control group}$
\end{center}
\vspace{.3cm}
+ $\overline{\textrm{\textit{voted}}}_\text{treatment group}$: proportion of registered voters who voted among those who received the message
\vspace{.2cm}
+ $\overline{\textrm{\textit{voted}}}_\text{control group}$: proportion of registered voters who voted among those who did not receive the message

## In-Class Exercise
1. Open RStudio (RStudio will open R)
\vspace{.2cm}

2. Open **inclass01.R** from within RStudio. Find it on Canvas > Modules.
    + RStudio: File >> Open File 

\vspace{.2cm}

3. Let's Run the code step-by-step.

## Does Social Pressure Affect Turnout?

```{r}
## STEP 1. Load the dataset
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
```

```{r}
## STEP 2. Look at the data
head(voting, 3) # shows the first three observations
## what does each observation represent?
## what is the outcome variable?
## what is the treatment variable?
```

## STEP 3. Create a binary treatment variable
- First, we need to learn how to use \fu{==} and \fu{ifelse()}
\vspace{.2cm}

- The operator \fu{==} \ tests whether the observations of a variable are equal to a particular value
    + values should be in quotes if text but without quotes, if numbers
    + examples: 
        + \fu{\textit{data}\$\textit{variable}==1} 
        + \fu{\textit{data}\$\textit{variable}=="yes"}
        
## STEP 3. Create a binary treatment variable

The function \fu{ifelse()}
\vspace{.2cm}

- creates the contents of a new variable based on the values of an existing one

- requires three arguments, separated by commas, in the following order: \newline (1) logical test (using \fu{==}), \newline (2) return value if logical test is true, \newline (3) return value if the logical test is false

\vspace{.2cm}

- Example: \fu{ifelse(\textit{data}\$\textit{variable}=="yes", 1, 0)} 
    
## STEP 3. Create a binary treatment variable

```{r}
## STEP 3. Create a binary treatment variable
## create variable pressure inside dataframe voting
voting$pressure <- # stores return values in new variable
  ifelse(voting$message=="yes", # logical test
         1, # return value if the logical test is true
         0) # return value if the logical test is false
```

- You should run the code all at once (not line by line)

## STEP 3. Create a binary treatment variable
- Whenever we create a new variable, we should make sure it was created correctly by looking at it:

```{r, myftnsize = TRUE}
head(voting, 4) # shows first observations
```

- Note that when *message* equals "yes", *pressure* equals 1; and when *message* equals "no", *pressure* equals 0

## STEP 4. Compute the difference-in-means estimator
\vspace{.3cm}
\begin{mdframed}[linecolor=gray, linewidth=0.05cm]
\vspace{.1cm}
\begin{center}
\color{concept}$\overline{Y}_\text{treatment group} - \overline{Y}_\text{control group}$\\[.3cm]
\color{gray}
$\overline{Y}_\text{treatment group}$: average outcome for the treatment group\\
$\overline{Y}_\text{control group}$: average outcome for the control group
\end{center}
\end{mdframed}
\vspace{.2cm}

- In the voting experiment:
\vspace{.2cm}
\begin{center}
$\overline{\textrm{\textit{voted}}}_\text{treatment group} - \overline{\textrm{\textit{voted}}}_\text{control group}$
\end{center}
\vspace{.3cm}
+ $\overline{\textrm{\textit{voted}}}_\text{treatment group}$: mean of *voted* for treatment group
+ $\overline{\textrm{\textit{voted}}}_\text{control group}$: mean of *voted* for control group
\vspace{-1cm}
        
## STEP 4. Compute the difference-in-means estimator
- Let's start by practicing computing and interpreting means

```{r}
mean(voting$voted) # calculates the mean of voted
```
- Interpretation?
    + 31\% of *all* the registered voters who were part of the experiment voted
- Why in \%?
    + Because *voted* is binary

## STEP 4. Compute the difference-in-means estimator
- \fu{mean(voting\$voted)} computes the mean of *voted* for all the observations in the dataset
- To compute the difference-in-means estimator, we need to calculate the mean of *voted* for subsets of observations
- Specifically, we need to compute:
    + the mean of *voted* for the treatment group (for which *pressure* equals 1)
    + the mean of *voted* for the control group (for which *pressure* equals 0)
\vspace{.2cm}
- To do this, we need to learn how to use the \fu{[]} \ operator

## STEP 4. Compute the difference-in-means estimator
- Operator \fu{[]}:
\vspace{.1cm}
    + extracts a selection of observations from a variable
    + to its left, we specify the variable we want to subset
    + inside the square brackets, we specify the criteria of selection
    + example: \fu{\textit{data}\$\textit{var1}[\textit{data}\$\textit{var2}==1]} \newline \color{gray}\# extracts the observations of the variable \textit{var1} for which the variable \textit{var2} equals 1
\vspace{-1cm}

## STEP 4. Compute the difference-in-means estimator

```{r, collapse=TRUE}
mean(voting$voted[voting$pressure == 1]) # treatment
mean(voting$voted[voting$pressure == 0]) # control
```
- Interpretation of the first mean?
    + 38\% of the registered voters who received the message voted (38x100=38\%)

\vspace{0.2cm}

- Interpretation of the second mean?
    + 30\% of the registered voters who did *not* receive the message voted (30x100=30\%)

## STEP 4. Compute the difference-in-means estimator

```{r, collapse=TRUE}
mean(voting$voted[voting$pressure==1]) - 
  mean(voting$voted[voting$pressure==0]) 
```

What are the effect's direction, size, and unit of measurement?

## STEP 5. Write conclusion statement
- What assumptions are we making when estimating the average causal effect? 
    + registered voters who received the message are comparable to registered voters who did not

\vspace{0.2cm}

- Why is this a reasonable assumption?
    + data come from a randomized experiment


## STEP 5. Write conclusion statement
- What's the treatment? 
    + receiving the message inducing social pressure

\vspace{0.2cm}

- What's the outcome? 
    + probability of voting

\vspace{0.2cm}

- What's the average causal effect's direction, size, and unit of measurement?
    + an increase of 8 percentage points, on average


## STEP 5. Write conclusion statement
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.2cm}
\begin{center}\concept{\tip{conclusion statement}}\end{center}

Assuming that \color{gray}[the treatment and control groups are comparable] \color{black} (a reasonable assumption because \color{gray}...\color{black}), \newline we estimate that \color{gray}[the treatment] [increases/decreases] \newline [the outcome] \color{black} by \color{gray} [size and unit of measurement of the effect]\color{black}, on average.
\end{mdframed}
\vspace{.2cm}
 
- Assuming that \pause registered voters who received the message are comparable to the registered voters who did not \pause (a reasonable assumption because \pause the data come from a randomized experiment), \pause we estimate that \pause receiving the message inducing social pressure \pause increases \pause the probability of voting by \pause 8 \pause percentage points, \pause on average.

## Summary

* **Today's Class:**
  + Units of Measurement of Means and Diffs-in-Means
  + In-Class Exercise: Does Social Pressure Increase the Probability of Turning Out To Vote?

* Next class:
  + Surveys
  
# Questions?

# See you in the next class!