---
title: |
    | **POLI 30 D: Political Inquiry**
    | \vspace{-.8cm}
author: |
    | \large\textbf{TA Sessions}\vspace{.3cm}
date: |
    | \large\textbf{Lab 05 | R Plots and R Data Analysis I}
fontsize: 12pt
urlcolor: blue
output:
  beamer_presentation:
    highlight: tango
    includes:
      in_header: '../../latex/style.tex'
    incremental: true # if true: pauses are inserted after each item in a list
    keep_tex: false # if true: .tex files are kept (helpful when fixing errors)
    slide_level: 2
classoption: "handout" # overrides incremental, removes all pauses
---

```{r, echo=FALSE, results='hide', include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysmall = function(before, options, envir) { 
  if (before) { 
    return("\\small") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(myfntsize = function(before, options, envir) { 
  if (before) { 
    return("\\footnotesize") 
  } else { 
    return("\\normalsize") 
  } 
}) 
knitr::knit_hooks$set(mytiny = function(before, options, envir) { 
  if (before) { 
    return("\\tiny") 
  } else { 
    return("\\normalsize") 
  } 
})
knitr::opts_chunk$set(mysmall = TRUE)
library(tidyverse)
```

## Before we start

**Announcements:**

- GitHub page: [**https://github.com/umbertomig/POLI30Dpublic**](https://github.com/umbertomig/POLI30Dpublic)

- Piazza forum: The link in the slides needs to be fixed. Check with instructors for an alternative link.

## Before we start

**Recap:** In the Lab sessions, you learned:

- How to install R and R Studio on your computer.
- How to do basic math operations in R.
- How to do basic vector and data.frame operations in R.
- How to install packages and work with R Markdown.

**Great job!**

- Do you have any questions about these contents?

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Plan for Lab 05}}}\normalsize\\\\
- Factors\\
- Booleans and logical tests\\
- Lists\\
- Tables\\
- Intro to plots\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}

# Factors

## Factors

- There is another type of variable that you learned little about: `factor`.

- Factors are great when you want to build tables because it takes a number and adds a mask to it.

- Raw content:

```{r, collapse = TRUE}
voted <- c(0,0,1,0,1) # 0=no and 1=yes
table(voted)
```

## Factors

- Factor:

```{r, collapse = TRUE}
voted_factor <- factor(voted, 
                       levels = c(0,1), 
                       labels = c('No', 'Yes'))
table(voted_factor)
```

## Factor

- The side-effect is that you cannot take means of it anymore, but you can still do `prop.table`s:

```{r, collapse = TRUE}
prop.table(table(voted_factor))
```

- And the mean would be equal to `r prop.table(table(voted_factor))[2]`.

# Booleans and Logics

## Booleans

- On programming, we frequently perform logical operations.

- For instance, suppose you have a vector of ages:

```{r, collapse = TRUE}
ages <- c(40, 20, 64, 20, 26, 21, 29, 32)
ages
```

## Booleans

- Which ages are above 30?

```{r, collapse = TRUE, myfntsize = TRUE}
ages
ages > 30
```

- Note the `TRUE` and `FALSE`. These are the so-called Boolean values.

## Booleans

- Which ages are equal to 20?

```{r, collapse = TRUE, myfntsize = TRUE}
ages
ages == 20
```

- Note the `==`. This is a **comparison operator**. For an assignment, you can use `<-` or `=`.

## Booleans

- What are the comparison operators?

| **operator** | **operation** |
|:---:|:---:|
| `>` | bigger than |
| `>=` | bigger than or equal |
| `<` | less than |
| `<=` | less than or equal |
| `==` | equal |
| `!=` | different (not equal) |
| `!` | not |
| `|` | or |
| `&` | and |

## Booleans

- And we can use the operators for subsetting:

```{r, collapse = TRUE, myfntsize = TRUE}
ages[!(ages == 20)] # same as ages[ages != 20]
```

- **Your turn**: Select ages above 25 and below 50. Hint: First, create the logical operation that selects those; then, use the logic operation to subset.

# Lists

## Lists

- All objects we have seen so far are of a fixed type.

- This makes it hard for you to store a text and a number in a placeholder without messing up with either the text or the number.

- Lists can hold any object without changing its type.

- But they are harder to operate.

## Lists

Example:

```{r, collapse = TRUE}
mylist <- list('numbers' = c(1,2,3), 
               'ages' = c(19,20,33), 
               'text' = 'My name is Tom')
mylist
```

## Lists

To index a list, do:

```{r, collapse = TRUE}
mylist[[1]]
mylist$text
```

And to add to a list:

```{r, collapse = TRUE}
mylist$newelement = 3
```

- We will not use them much, but it is nice that you know they exist.

# Tables

## Tables

- We know you already saw a few of those, but still, let's talk about them.

- Tables count the number of elements and display the label and the frequency of the elements.

- Let's load the voting dataset:

```{r, collapse = TRUE}
voting <- read.csv("https://raw.githubusercontent.com/umbertomig/POLI30Dpublic/main/data/voting.csv")
```

## Tables

- And a table of messages looks like this:

```{r, collapse = TRUE}
table(voting$message)
```

- The syntax is always like this: `table(dataframename$variablename)`.

- **Your turn**: Make a table of the variable `voted`.

## Tables

- If we want a proportions table, we can add `prop.table` to a table:

```{r, collapse = TRUE}
prop.table(table(voting$message))
```

- The syntax is always like this: `prop.table(table(dataframename$variablename))`.

- **Your turn**: Make a proportions table of the variable `voted`.

- **Your turn**: Make a proportions table of the variable `birth`. Does it look good?

# Intro to Plots

## Plots of a single variable

- For plots, we will use a package called `ggplot2`.

- Here is a good [cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf). This is a great thing to print and has close by when creating plots.

- `ggplot2` is based on the `grammar of graphs`. But what is this?

## Plots of a single variable

- In abstract, the grammar of graphs is a decomposition of plots in its main features.

- In essence, every plot has:

1. A dataset
2. A coordinated system
3. A geometric shape

- And different plots are different compositions of these three key ingredients.

## Plots of a single variable

- Histograms are great plots for continuous variables. The syntax is:

```
ggplot(data = dataset, aes(x = variable_x_name)) +
  geom_histogram()
```

- Note the components in action: 
  + `data` adds the dataset; 
  + `aes(x =...)` adds the coordinated system; 
  + `geom_histogram()` adds the geometric shape.

## Plots of a single variable

```{r, fig.height=2, fig.width=2, fig.align='center', mytiny = TRUE}
ggplot(data = voting, aes(x = birth)) + geom_histogram()
```

\vspace{-2cm}

## Plots of a single variable

- And we can customize it a lot. For the `geom_histogram()`:
  + `stat =`: Default is `bin` 
  + `bins = `: Number of bins (default is 30)
  + `binwidth = `: Size of the bin (don't use when using `bins`)
  + `breaks = `: Vector with custom breaks (overrides `bins` and `binwidth`)
  + `color = `: Changes the contour colors.
  + `fill = `: Changes the color of the shape.
  + `alpha = `: Changes the transparency (between 0 and 1, with 0 being transparent).
  + `aes(y = ..density..)`: Instead of count, show densities (good when we have large datasets).

\vspace{-1cm}

## Plots of a single variable

- For the main plot:

```
plot + 
  labs(x = 'X-axis title',
       y = 'y-axis title',
       title = 'plot title',
       subtitle = 'subtitle below the plot')
```

## Plots of a single variable

- And the `theme` changes the general look of the plot:
  + `theme_bw()`: Black and white. My personal choice.
  + `theme_gray()`: Gray background.
  + `theme_dark()`: Dark gray background.
  + `theme_minimal()`: Little structure.
  + `theme_void()`: Empty theme.

- Now, let's see a nice plot with lots of customization.

## Plots of a single variable

```{r, fig.height=2.5, fig.width=2.5, fig.align='center', mytiny = TRUE}
ggplot(data = voting, aes(x = birth)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, color = 'black', fill = 'lightblue', alpha = 0.6) +
  labs(title = 'Histogram of Birth Year', x = 'Birth Year', y = '') + theme_minimal()
```

\vspace{-2cm}

## Plots of a single variable

- And for a continuous variable, instead of `geom_histogram`, you could do:
  + `geom_area(stat = 'bin')`: Looks like a big polygon
  + `geom_density()`: Density plot
  + `geom_dotplot()`: Very cool plot with little dots instead of rectangles.
  + `geom_line()`: Line-plot: pretty much connects things with lines.
  + `geom_step()`: Very cool, like a contour of a histogram.
  
- Try them out and see. Some would need some customizing before it works.

\vspace{-1cm}

***
\begin{center}
\begin{minipage}{0.6\paperwidth}
\begin{mdframed}[backgroundcolor=white,linecolor=title, linewidth=0.05cm]
\vspace{.5cm}
\begin{center}
\begin{tabular}{l}
\multicolumn{1}{c}{\large\concept{\textbf{Today's Lab}}}\normalsize\\
- Factors\\
- Booleans and logical tests\\
- Lists\\
- Tables\\
- Intro to plots\\\\
\multicolumn{1}{c}{\large\concept{\textbf{Next Lab}}}\normalsize\\
- More plots\\
- Some less obvious data analysis\\\\
\end{tabular}
\end{center}
\vspace{.5cm}
\end{mdframed}
\end{minipage}
\end{center}
  
# Questions?

# See you in the next lab!